\appendix
\chapter{Reproduction and Artifact Manifest}

The repository root contains \texttt{REPRODUCING.md}, \texttt{evidence-manifest.json}, the locked dependency graph, Docker configurations, and the versioned corpus. Legacy reports are explicitly excluded. A final accepted manifest must contain the clean Git commit, Docker digests, corpus hash, run identifiers, and hashes of generated thesis tables.

\section{Command Sequence}

\begin{Verbatim}[breaklines=true,fontsize=\small]
uv sync --locked --group dev
uv run ruff check src tests scripts
uv run mypy src/unified_multi_agent_coordination
uv run pytest --cov --cov-branch --cov-report=json
uv run python scripts/check_coverage.py coverage.json
uv run unified-coordination-scenarios --no-output
$env:EVIDENCE_RUN_DIR="demo_runs/system/<new-run-id>"
docker compose -f docker-compose.system.yml up --build \
  --abort-on-container-exit --exit-code-from system-tests
$env:EVIDENCE_RUN_DIR="demo_runs/postgres/<new-run-id>"
docker compose -f docker-compose.distributed.yml up --build \
  --abort-on-container-exit --exit-code-from distributed-system-tests
uv run unified-consensus-matrix \
  --output-dir demo_runs/consensus/<new-run-id> --trials 3 \
  --promotion-candidate
uv run unified-defense-study --check
uv run unified-analyze-defense-study \
  demo_runs/v0.3/20260713T013624Z-b9fd8b6f83 \
  --corpus corpus/v0.3 \
  --output demo_runs/v0.3/20260713T013624Z-b9fd8b6f83/analysis.json
uv run unified-defense-study-v04 --corpus corpus/v0.4 --check
uv run unified-analyze-defense-study-v04 \
  --run demo_runs/v0.4/20260713T121506Z-a0ecd2bf0e \
  --corpus corpus/v0.4 \
  --output demo_runs/v0.4/20260713T121506Z-a0ecd2bf0e/<new-analysis>.json
uv run unified-runtime-ablation-study \
  --output-dir demo_runs/runtime_ablation/<new-run-id>
uv run unified-evidence-preflight --require-release \
  --corpus corpus/v0.5 --repository .
\end{Verbatim}

\chapter{Predicate Refinement and Claim--Evidence Matrix}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.22\textwidth}>{\raggedright\arraybackslash}p{0.35\textwidth}>{\raggedright\arraybackslash}p{0.33\textwidth}}
\toprule
\textbf{Claim or predicate} & \textbf{Implementation refinement} & \textbf{Evidence required} \\
\midrule
Coverage & Canonical capability ID, full modes and schema coverage & Feasibility and ID-perturbation tests \\
Authority & Local trust order and admission policy & Remote-card and credential tests \\
Constraints & Typed source, JSON pointer, operator and expected value & Missing, type, false and unresolved cases \\
Verifiability & Task, requirement or capability validation contract & Validatorless and nested-schema tests \\
Dependency safety & Completed prerequisites and declared artifacts only & Failure/timeout/unknown branch tests \\
Traceability & Persisted session, plan, task and attempt events & Ordering, restart and isolation tests \\
Recovery & Lease, fencing, idempotency and unknown-side-effect policy & PostgreSQL crash-fault harness \\
Distributed authority & etcd/Raft state, learner-first membership, generation CAS, resumable phases, and quorum fail-closed behavior & Historical three-voter report plus required clean 3/5/7 campaign \\
RQ4 comparison & Fresh author-labelled, label-hidden v0.5 matched-pair study & Exact 1,440-output dual-arm matrix, deterministic analysis, and reproducible digest \\
Runtime mechanisms & Secure-default policy surfaces and explicit negative controls & Separate 350-observation runtime-only bundle \\
\bottomrule
\end{longtable}

\section{Research-Question Traceability}

\begingroup
\setlength{\tabcolsep}{3pt}
\begin{longtable}{>{\raggedright\arraybackslash}p{0.08\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}>{\raggedright\arraybackslash}p{0.28\textwidth}}
\toprule
\textbf{RQ} & \textbf{Claim} & \textbf{Mechanism and evidence} & \textbf{Current boundary} \\
\midrule
RQ1 & Executability requires explicit identity, coverage, authority, dependencies, artifacts, validation, and completion. & Typed models, feasibility predicates, hydration, artifact and trace tests. & Prototype predicates; no proof for arbitrary domains. \\
RQ2 & Linguistic proposal can be separated from deterministic authorization. & Non-authoritative drafts, symbolic gate, authorization-before-dispatch tests, and four v0.5 analysis views. & Implemented separation; author-labelled cases and small local models. \\
RQ3 & Temporary language capabilities can be bounded and auditable. & Task-local auxiliary specification, admission and expiry tests, separate runtime ablation. & Read-only fixture scope; no hostile generated code. \\
RQ4 & The constrained bridge is compared fairly with a direct LLM coordinator. & Complete 1,440-output v0.5 matrix, equal public inputs and call budgets, McNemar tests, and pair-cluster bootstrap. & Safe refusal supported; useful acceptance and superiority unsupported; no independent labels. \\
RQ5 & A replicated coordinator should preserve one authorization state under measured crash and reconfiguration faults while a majority survives. & etcd store, monotonic fences, durable membership state machine, quorum error contract, complete 33-trial campaign. & Unsupported: 20 passed, seven invariant failures, and six infrastructure errors on one host. \\
\bottomrule
\end{longtable}
\endgroup

\chapter{Tribunal Questions and Defensible Boundaries}

\begin{description}
  \item[What is novel?] The protocol-adaptable integration and operational refinement of a linguistic proposal boundary, symbolic authorization/refusal, durable evidence, admission policy, and measured recovery. Individual mechanisms are prior art.
  \item[Why use an LLM?] Natural-language interpretation and candidate decomposition vary semantically; a structured rule oracle assumes that work is already done. The LLM remains non-authoritative.
  \item[What is guaranteed?] The tested implementation enforces declared predicates and ordering under its trusted-code and trusted-store assumptions. It does not guarantee semantic truth or runtime success.
  \item[What is trusted?] The coordinator code, durable store, locally configured trust policy, admitted registry descriptions, and contract validators. Malicious admitted agents remain outside scope.
  \item[Is fencing consensus?] Fencing alone is not consensus. The legacy PostgreSQL mode serializes through one database; the etcd mode separately uses Raft consensus to order authoritative state. Fencing then prevents stale session owners from committing and must also be enforced by effectful receivers.
  \item[Does the cluster scale automatically?] Membership of already running coordinator nodes is discovered and reconciled automatically. The software does not provision compute, and permanent majority loss requires operator-led disaster recovery.
  \item[Is there no single point of failure?] With at least three voters placed in independent failure domains, no single coordinator or etcd-member crash stops the authoritative path. One voter, one Docker host, shared network infrastructure, and a cluster-wide secret remain single-failure or correlated-failure risks.
  \item[How broad is A2A interoperability?] Controlled fixtures plus one vendored, upstream-derived sample pinned to an upstream commit. This is evidence of concrete compatibility, not universal conformance.
  \item[Did the hybrid work empirically?] Only as a fail-closed boundary. It made zero unsafe acceptances but accepted five of 360 feasible repeated observations after repair. It was neither useful nor more accurate than the direct arm on this corpus.
  \item[Why trust the comparison?] The corpus and author labels were frozen before collection, labels were hidden until both arms completed, all 1,440 identities and equal call budgets were verified, and independent analysis runs reproduced the same digest. Labels were not independently adjudicated, so claims remain limited to framework conformance on this designed corpus.
  \item[Is RQ5 answered?] Yes, negatively. The clean campaign was complete and valid, but only 20 of 33 trials passed; seven violated invariants and six had infrastructure errors. The measured implementation does not support the full RQ5 claim.
\end{description}

\chapter{Disclosure of Artificial Intelligence Use}
\label{app:ai-disclosure}

Artificial-intelligence systems were used in two distinct roles during this thesis: as experimental subjects whose outputs form part of the evaluation, and as assistive tools used during research, software development, and manuscript revision. These roles must not be conflated. Experimental model outputs are immutable study observations; assistive outputs are suggestions that remain subject to author review. In every role, responsibility for the research design, reference labels, source selection, implementation, interpretation, claims, and submitted text remains with the author.

\clearpage
\section{Systems, Tools, and Purposes}

\begingroup
\setlength{\tabcolsep}{4pt}
\begin{longtable}{>{\raggedright\arraybackslash}p{0.19\textwidth}>{\raggedright\arraybackslash}p{0.28\textwidth}>{\raggedright\arraybackslash}p{0.43\textwidth}}
\toprule
\textbf{System or tool} & \textbf{Role} & \textbf{Use in the thesis} \\
\midrule
\endfirsthead
\toprule
\textbf{System or tool} & \textbf{Role} & \textbf{Use in the thesis} \\
\midrule
\endhead
OpenAI Codex (GPT-5-based project sessions) & Interactive research, writing, and software-development assistant & Used to critique chapter structure and claims; assist with literature discovery and source triage; draft or revise prose, documentation, and the bilingual abstract; propose code, refactorings, tests, evaluation scripts, CI checks, and reproducibility instructions; inspect claim--evidence consistency; and assist with LaTeX/PDF quality review. Suggested citations were not accepted as evidence without inspection of the underlying source. All changes remained visible in the repository and were reviewed against tests, static checks, evidence files, or the compiled manuscript. \\
LM Studio & Local model-serving environment & Used to load and serve the pinned local models through an OpenAI-compatible HTTP interface for the v0.3--v0.5 collection protocols. It provided the execution environment for study inference; it was not treated as an author, adjudicator, or source of scientific claims. \\
Qwen3 1.7B, Gemma 4 E2B, and Qwen3 8B & Experimental language models & Used as study subjects to produce non-authoritative linguistic drafts or direct coordination decisions under frozen prompts, schemas, cases, arms, and requested seeds. Raw and repaired responses were preserved and scored only after collection completion. These models did not receive hidden labels and were not used as independent validators of those labels. \\
Lingo & Typed linguistic-application framework & Used inside the prototype to expose bounded interpretation, extraction, classification, and candidate-planning capabilities through explicit state and schemas. Lingo is implementation middleware rather than the source of the evaluated model weights or an authority over coordination decisions. \\
OpenAI-compatible client interfaces & Model-transport interface & Used to send local requests to LM Studio and record responses, metadata, latency, and errors. Use of the interface does not imply that hosted OpenAI models generated the experimental observations. \\
\bottomrule
\end{longtable}
\endgroup

AI assistance also contributed to exploratory scenario, prompt, baseline, and test-design critique. The author selected the final protocol, determined and froze the reference labels, and accepted responsibility for their correctness. AI assistance is therefore not presented as independent annotation or adjudication; the absence of an independent human adjudicator remains a threat to validity.

\section{Authorial Control and Verification}

AI-generated or AI-revised material was not accepted solely because it was fluent or technically plausible. Manuscript changes were checked against primary sources and repository evidence. Code changes were inspected and exercised through the applicable Ruff, mypy, pytest, coverage, deterministic, container, and evidence-preflight checks. Numerical statements were derived from immutable JSON evidence or generated LaTeX macros. Statistical analyses were implemented as deterministic programs and reproduced by digest. Figures in the submitted manuscript are repository-authored LaTeX/TikZ artifacts; no generative image system was used for the current figures.

The AI assistants did not determine the final research questions, authorize execution in the evaluated system, alter collected model outputs, fabricate measurements, select favorable runs, assign themselves authorship, or replace the author's interpretation of negative outcomes. Working conversations with an assistant may contain provisional ideas and rejected text; they are not research evidence. The version-controlled manuscript, source code, frozen protocols, raw outputs, analysis programs, manifests, and accepted reports are the auditable artifacts.

For clarity, the principal non-AI tools used to construct and validate the artifact include Python and \texttt{uv}; Pydantic, FastAPI, HTTPX, and Uvicorn; pytest, coverage.py, Ruff, and mypy; Git and GitHub Actions; Docker Compose; PostgreSQL; etcd/Raft; the A2A SDK and upstream-derived sample; JSON Schema; and LuaLaTeX, BibTeX, TikZ, Poppler, and PDF inspection tools. These tools execute deterministic software, infrastructure, validation, or document-processing functions and are not reported as generative-AI authorship assistance.

No other generative-AI assistant is documented in the project record. If an additional assistant was used outside that record, it must be added to this disclosure before submission. This statement is intended to make the role of AI inspectable without transferring scholarly responsibility away from the author.
