\chapter{State of the Art}
\label{chap:state-of-the-art}

\section{Framing the Research Problem}

The contemporary discussion of artificial intelligence agents is marked by a productive but sometimes imprecise convergence between two research traditions. The first tradition is the classical computer-science study of autonomous agents and multi-agent systems, in which an agent is a situated computational entity that perceives an environment, reasons over goals or policies, and acts through effectors \cite{wooldridge1995intelligent,russell2021artificial,jennings2000agent}. The second tradition is the recent development of large language model (LLM) applications that use natural language, tool invocation, memory, and planning prompts to produce behavior that is described as agentic \cite{brown2020language,wei2022chain,yao2023react,wang2024survey}. The overlap between these traditions is substantial, but not identical. Classical agents emphasize autonomy, reactivity, proactiveness, social ability, commitments, and environment interaction; LLM agents often emphasize linguistic reasoning, flexible task decomposition, and interaction with external tools.

This distinction is central to a thesis on unified multi-agent coordination. A chatbot that produces a response to a prompt is not, by itself, an autonomous agent in the classical sense. Conversely, a symbolic controller that follows a plan can be an agent even if it has no natural-language interface. The research opportunity appears when LLMs are treated as components inside broader agent architectures rather than as replacements for agency itself. In that view, an LLM can serve as a linguistic reasoning module, a translator between human instructions and formal actions, an explanation generator, or a coordinator of high-level intent. The agent, however, remains the larger computational system that maintains state, chooses actions, interacts with other agents, and is accountable for execution.

The literature therefore motivates a bridging problem. Classical multi-agent systems (MAS) provide mature formalisms for interaction protocols, negotiation, task allocation, organizational roles, commitments, and verification \cite{wooldridge2009introduction,weiss2013multiagent,ferber1999multi,singh1999commitments}. LLM-based systems provide flexible natural-language mediation, few-shot generalization, and tool-use interfaces \cite{ouyang2022training,schick2023toolformer,mialon2023augmented}. Symbolic AI supplies explicit representations for planning, constraint satisfaction, rule-based control, and reasoning over dynamic worlds \cite{ghallab2004automated,reiter2001knowledge,hogan2021knowledge}. A unified coordination architecture must integrate these strengths without collapsing their differences. In particular, it must support agents that communicate linguistically, reason symbolically, invoke tools safely, and participate in multi-agent coordination mechanisms whose semantics are explicit enough to be analyzed.

The phrase state of the art is used here in a precise sense. It does not refer only to the most recent LLM frameworks or protocols, because novelty alone is not sufficient for autonomous coordination. Instead, the relevant state of the art consists of the strongest available concepts, formalisms, systems, and evaluation practices that inform the design of a coordinated autonomous system. This includes older but still foundational work on agency and communication, mature symbolic methods for action and verification, and newer LLM-based systems that demonstrate flexible linguistic control. The purpose of the chapter is therefore synthetic: it identifies what each research line contributes, where each line remains insufficient, and why a unified LLM-symbolic MAS architecture is a coherent research target.

\section{Classical Foundations of Autonomous Agents}

The classical agent literature provides the conceptual baseline for this thesis. Wooldridge and Jennings characterize agents through autonomy, social ability, reactivity, and proactiveness, thereby distinguishing agents from passive objects or services \cite{wooldridge1995intelligent}. Russell and Norvig formulate the rational-agent view, in which an agent maps percept histories to actions and is evaluated by a performance measure in an environment \cite{russell2021artificial}. These definitions remain important because they locate agency in a control loop: perception, internal state update, decision, and action. The same loop is required for autonomous systems whether their reasoning module is symbolic, neural, or hybrid.

Several architectural families have shaped the design of autonomous agents. Reactive agents connect percepts to actions through condition-action rules or layered behaviors. Brooks's subsumption architecture demonstrated that robust situated behavior could arise from distributed reactive layers without a centralized symbolic world model \cite{brooks1986robust}. Deliberative agents, by contrast, maintain explicit models and use search or planning to choose actions. Hybrid architectures combine these approaches, usually by placing a fast reactive layer under a slower deliberative layer. Cognitive architectures such as Belief-Desire-Intention (BDI) systems add explicit mental attitudes, representing what the agent believes, wants, and has committed to doing \cite{bratman1987intention,rao1991modeling,rao1995bdi,georgeff1999belief}.

BDI architectures are particularly relevant to the present thesis because they provide a disciplined vocabulary for autonomous behavior. Beliefs encode information about the world, desires encode possible objectives, and intentions encode commitments to plans. This separation is useful for LLM-symbolic systems because LLM output should not automatically become a commitment to act. A linguistic proposal may be treated as a candidate desire, an explanation, or a plan sketch; the symbolic or procedural layer can then validate whether it should become an intention. This division helps address a common weakness of LLM agents: fluent text can appear decisive even when the underlying action has not been checked against state, preconditions, or policy.

Agent-oriented programming also contributes important design principles. Shoham proposed programming in terms of mentalistic concepts such as beliefs, obligations, and capabilities \cite{shoham1993agent}. Later platforms such as JADE, Jason, GOAL, and JaCaMo operationalized agent communication, BDI reasoning, and multi-agent organization in software frameworks \cite{bellifemine2007developing,bordini2007programming,hindriks2009programming,boissier2013multi}. These systems show that agent design benefits from explicit abstractions for messages, roles, plans, and norms. Modern LLM frameworks often rediscover these abstractions through role prompts, tool registries, memory modules, and conversation controllers, but the classical literature provides more precise semantics for them.

The boundary between agent and non-agent software is therefore not merely terminological. A database, an API, a workflow, and a model endpoint can all participate in an agentic system, but they do not necessarily exhibit agency. An agent must have a decision procedure and some degree of autonomy with respect to its environment \cite{franklin1997agent}. This matters for system design because different components require different forms of assurance. A tool can be validated by its interface contract; a model can be evaluated by prediction or generation quality; an autonomous agent must also be evaluated by its policy for choosing when and how to act. In LLM-based systems, this distinction is often obscured when every tool wrapper or prompt template is called an agent. A rigorous thesis architecture should reserve the term agent for components that participate in a sense-reason-act loop with explicit state and responsibility.

\begin{table}[htbp]
\centering
\caption{Agent architectural traditions relevant to unified coordination.}
\label{tab:agent-architectures}
\begin{tabular}{p{0.24\textwidth}p{0.66\textwidth}}
\toprule
\textbf{Tradition} & \textbf{Relevance to this thesis} \\
\midrule
Reactive agents & Provide fast environment response and motivate safety shields that can override slower deliberation. \\
Deliberative agents & Provide explicit planning, goal reasoning, and action selection over symbolic state. \\
BDI agents & Separate beliefs, desires, and intentions, which is useful for interpreting LLM proposals before execution. \\
Hybrid agents & Combine reactive control with deliberative reasoning, matching the need to combine LLM reasoning with symbolic execution. \\
Agent-oriented programming & Provides software abstractions for roles, commitments, messages, plans, and organizational structure. \\
\bottomrule
\end{tabular}
\end{table}

\section{Multi-Agent Coordination}

Multi-agent systems extend the agent concept from individual autonomy to distributed computation among multiple autonomous entities. In a MAS, agents may cooperate, compete, negotiate, delegate, or coordinate around shared resources \cite{ferber1999multi,weiss2013multiagent,wooldridge2009introduction}. The key difficulty is that no single agent necessarily possesses complete knowledge, global control, or unlimited authority. Coordination is therefore required to manage dependencies among activities, allocate tasks, resolve conflicts, synchronize plans, and maintain coherent collective behavior. Malone and Crowston define coordination broadly as the management of dependencies between activities \cite{malone1994interdisciplinary}; the MAS literature operationalizes this idea through protocols, organizational structures, distributed algorithms, and shared environments \cite{jennings1996coordination}.

Task allocation is one of the oldest and most concrete coordination problems. The Contract Net Protocol introduced a manager-contractor pattern in which tasks are announced, agents bid, and contracts are awarded according to suitability or cost \cite{smith1980contract}. Market-based coordination generalizes this approach by using auctions, marginal costs, and economic mechanisms to allocate work or resources \cite{sandholm1993implementation,wellman1993market}. Coalition formation addresses cases in which a task requires the combined capabilities of multiple agents \cite{shehory1998methods}. These mechanisms are directly relevant to LLM-based multi-agent systems, where role-specialized agents may be dynamically selected for subtasks such as planning, coding, verification, retrieval, or simulation.

Shared-state coordination offers a different model. Blackboard systems allow specialized knowledge sources to read and write partial solutions in a common workspace \cite{nii1986blackboard,corkill1991blackboard}. Tuple-space coordination, exemplified by Linda, decouples agents through generative communication over shared tuples \cite{gelernter1985generative}. Stigmergic coordination uses changes in the environment as indirect signals, a concept developed in studies of social insects and later applied to swarm systems \cite{theraulaz1999brief}. These mechanisms anticipate modern agent workspaces, shared memories, vector stores, and artifact repositories. They also expose a risk: if the shared state is ambiguous or corrupted, coordination can fail even when each individual agent appears locally competent.

Distributed constraint reasoning provides a more formal approach to joint decision-making. Distributed Constraint Satisfaction Problems and Distributed Constraint Optimization Problems model decisions as variables controlled by different agents under local and global constraints \cite{yokoo1998distributed,modi2005adopt,fioretto2018distributed}. Such formalisms are valuable for autonomous systems because they make resource limits, task dependencies, and optimization criteria explicit. LLM systems rarely use DCOP-style models directly, but a unified architecture can use LLMs to elicit goals or preferences while a symbolic optimizer enforces constraints. This division is especially important when natural-language instructions are incomplete, inconsistent, or underspecified.

Social and normative approaches broaden coordination beyond efficiency. Commitments, norms, institutions, trust, and argumentation provide mechanisms for regulating interactions among autonomous agents \cite{singh1999commitments,dignum2004model,rahwan2009argumentation}. In classical MAS, a commitment can represent an obligation from one agent to another; protocols can then be evaluated in terms of how commitments are created, discharged, or violated. This perspective is relevant to LLM agents because dialogue alone does not guarantee accountability. If one agent asks another to perform an action, a unified system should represent whether that request created an obligation, whether the target accepted it, and what evidence establishes completion.

Coordination failure is as important as coordination success. Distributed systems may fail through deadlock, livelock, inconsistent local beliefs, resource contention, duplicated effort, or incompatible goals. LLM-based MAS add further failure modes: agents may agree linguistically while using different internal interpretations; a critic agent may approve an artifact without checking it; multiple agents may reinforce the same hallucination; or a delegated task may be considered complete because the conversation ended rather than because a verifiable postcondition was reached. Classical MAS mechanisms provide vocabulary for diagnosing these failures. For example, a protocol trace can reveal that an expected acceptance message was never sent, and a commitment record can reveal that a promised action lacks evidence of completion. This is a strong argument for storing coordination state in explicit symbolic structures rather than only in chat history.

The distinction between centralized and decentralized coordination also requires careful treatment. Centralized orchestration simplifies global control and is common in practical LLM systems, where a controller routes messages among role-specific agents. Decentralized coordination can improve robustness and scalability, but it requires stronger assumptions about local decision-making and communication semantics. Hybrid coordination is often more realistic: a central orchestrator may allocate tasks, while agents negotiate details, update shared artifacts, and call local tools. For autonomous systems, the correct choice is domain-dependent. Safety-critical execution may require centralized authorization, while exploratory planning may benefit from decentralized proposal generation and critique.

\section{Agent Communication Languages and Interaction Protocols}

Coordination depends on communication, but not all communication has the same semantics. Classical Agent Communication Languages (ACLs) were developed to provide structured, interoperable message exchange among agents. Their theoretical basis lies in speech-act theory, where utterances are understood as actions that can request, inform, promise, or commit \cite{austin1962how,searle1969speech,cohen1990intention}. KQML introduced performatives such as ask, tell, and reply, together with message fields for content, language, and ontology \cite{finin1994kqml,labrou1997proposal}. FIPA-ACL later standardized communicative acts and interaction protocols with more explicit semantics grounded in mental attitudes \cite{fipa2002acl}.

The importance of ACLs for this thesis is not that modern systems should necessarily reuse KQML or FIPA unchanged. Rather, the key lesson is that coordination messages require interpretable force. A statement such as "the route is clear" differs from "please clear the route" or "I commit to clearing the route." In natural language, such distinctions are often implicit. LLM agents can generate and interpret nuanced language, but they can also blur assertion, request, plan, and commitment. A unified coordination architecture should therefore map natural-language utterances to structured communicative acts when those utterances affect execution.

Interaction protocols provide the temporal structure of communication. Contract net, request-response, subscription, negotiation, and auction protocols specify allowed message sequences and expected state transitions \cite{fipa2002acl,jennings2001automated}. In classical MAS, this structure supports interoperability and verification. In LLM-based systems, conversation often proceeds through ad hoc chat transcripts. Such transcripts are flexible, but flexibility can make it difficult to determine whether a task has been accepted, delegated, blocked, or completed. A hybrid architecture can preserve natural-language flexibility for explanation and negotiation while retaining structured protocol state for execution.

Content languages and ontologies also remain central. Gruber defines an ontology as an explicit specification of a conceptualization \cite{gruber1993translation}; description logics provide formal foundations for class, relation, and constraint reasoning \cite{baader2003description}. Without shared semantics, agents may exchange syntactically valid messages that have different meanings to different participants. LLMs can help mediate heterogeneous vocabularies, but they cannot by themselves guarantee semantic alignment. For autonomous systems, especially those controlling tools or physical processes, semantic grounding must be explicit enough to support validation.

The transition from ACLs to natural-language coordination should therefore be understood as an expansion of expressive capacity rather than a replacement for structure. Natural language is valuable because it can express uncertainty, context, explanation, and negotiation without requiring all possibilities to be pre-enumerated. However, execution requires discrete commitments. An LLM may write, "I can handle the scheduling task," but a coordination layer must determine whether this is an assertion of capability, an offer, an acceptance, or a mere conversational signal. Classical ACLs solve this by using performatives; modern LLM systems can approximate the same effect by extracting structured message types, confidence, referenced artifacts, intended recipients, and expected follow-up actions. The extraction need not eliminate natural language; it should accompany it with machine-checkable coordination metadata.

Another issue is observability. In a free-form conversation, the state of the protocol is often implicit in the transcript. A human reader may infer that a task has moved from proposal to execution, but a verifier cannot safely depend on such inference. Structured communicative acts make protocol state observable. For example, a request can be represented with a sender, receiver, content, deadline, acceptance status, and completion criterion. This explicit state is compatible with LLM-generated messages, but it requires an architectural layer that treats dialogue as input to a coordination model rather than as the coordination model itself.

\section{Symbolic Foundations}

Symbolic AI provides the formal machinery needed to represent knowledge, define actions, and reason about consequences. Early work on logic and common-sense reasoning established the idea that intelligent behavior can be supported by symbolic descriptions of the world and action \cite{mccarthy1969some,newell1976computer}. Reiter's work on knowledge in action developed logical foundations for specifying dynamic systems \cite{reiter2001knowledge}. Knowledge graphs and ontologies extend this tradition by supporting semantic integration across heterogeneous data sources \cite{hogan2021knowledge}. In a unified agent architecture, these symbolic representations can serve as the stable substrate against which LLM-generated interpretations are checked.

Automated planning is especially relevant because it provides a disciplined account of goal-directed action. STRIPS introduced actions with preconditions and effects \cite{fikes1971strips}; PDDL later provided a standard planning-domain language that supported interoperability among planners \cite{mcdermott1998pddl,fox2003pddl2}. Planning systems search for action sequences that transform an initial state into a goal state under domain constraints \cite{ghallab2004automated}. This formalism directly addresses a weakness of LLM planning: language models can produce plausible plans without verifying action applicability, resource limits, or causal dependencies. A planner can reject or repair such plans when the formal model is available.

Hierarchical Task Network (HTN) planning adds a decomposition perspective. Rather than planning only through primitive actions, HTN methods refine abstract tasks into subtasks according to domain-specific methods \cite{erol1994htn,nau2003shop2,georgievski2015htn}. This is closely aligned with how LLM agents often decompose natural-language tasks into steps. The difference is that HTN decomposition is constrained by explicit methods, whereas LLM decomposition is often implicit in generated text. Combining the two can allow an LLM to propose a decomposition while a symbolic layer verifies that each refinement is permitted by the domain model.

Rule-based and non-monotonic reasoning also contribute to agent control. Logic programming and Answer Set Programming support declarative encodings of constraints, defaults, and alternative stable models \cite{gelfond1991classical,lifschitz2008answer}. These approaches are useful when coordination requires reasoning about exceptions, incomplete knowledge, or mutually exclusive choices. Finite-state machines, behavior trees, and Goal-Oriented Action Planning provide additional symbolic control structures that are widely used in robotics and games \cite{colledanchise2018behavior,orkin2006three}. Their advantage is not linguistic flexibility but predictable execution structure. In safety-critical autonomous systems, predictability is not optional.

Symbolic methods also clarify the difference between a plan and an explanation. LLMs can explain a possible course of action in coherent prose, but a plan in the symbolic sense is an object with preconditions, effects, ordering constraints, resources, and success criteria. Treating explanations as plans can lead to execution errors because the text may omit necessary conditions or represent them ambiguously. Conversely, symbolic plans can be difficult for humans to understand without explanation. A hybrid system can use a symbolic planner to produce or verify the executable structure while using an LLM to explain the plan, ask for missing information, or translate user intent into formal goals. The two artifacts should remain linked but distinct.

There are also limits to symbolic methods. A planner can only reason over what has been modeled. A rule engine can only enforce rules that are encoded. A knowledge graph may be incomplete or stale. The role of the LLM is therefore not only convenience; it can help interpret underspecified tasks, retrieve relevant context, and generate candidate formalizations. The strongest hybrid design is not a symbolic system with a superficial language interface, nor an LLM system with occasional tool calls. It is an architecture that assigns responsibilities according to epistemic strength: language models handle interpretation and communication; symbolic components handle explicit constraints, commitments, and execution semantics.

\section{Neuro-Symbolic AI and Hybrid Reasoning}

Neuro-symbolic AI studies the integration of neural learning with symbolic reasoning. Surveys emphasize that neural systems are effective for perception, pattern recognition, and language, while symbolic systems support abstraction, compositionality, explanation, and formal reasoning \cite{besold2017neural,garcez2019neural,hitzler2022neuro}. LLMs intensify this integration problem. They are neural models with broad linguistic competence, but their outputs are not inherently grounded in formal semantics. A hybrid architecture can exploit their strengths while delegating constraint satisfaction, planning, and verification to symbolic components.

One important pattern is LLM-to-symbolic translation. A language model may translate a user request into a PDDL goal, a program, a query, a constraint set, or an action schema. LLM+P, for example, explores the use of LLMs with classical planning by translating natural-language planning tasks into planner-compatible representations \cite{liu2023llmp}. Studies of LLM planning capabilities also show that LLMs are not reliable substitutes for planners when exact action semantics are required \cite{valmeekam2023planbench}. The implication is not that LLMs are unhelpful for planning; rather, they should be placed where linguistic flexibility is valuable and checked where formal correctness matters.

A second pattern is symbolic-to-LLM augmentation. Retrieval-augmented generation supplies external documents as context \cite{lewis2020retrieval}; tool-augmented systems expose calculators, search engines, databases, or APIs \cite{schick2023toolformer,mialon2023augmented}; MRKL systems propose modular architectures in which language models route tasks to specialized symbolic or computational modules \cite{karpas2022mrkl}. In agent systems, this pattern generalizes to planners, solvers, knowledge graphs, monitors, and execution kernels. The LLM becomes an interface and orchestrator, not the sole reasoner.

The most relevant pattern for this thesis is a hybrid execution loop. The agent perceives or receives a task, uses an LLM to interpret context and propose candidate actions, translates selected candidates into formal representations, uses symbolic modules to validate or refine them, executes through a controlled interface, and records feedback for future reasoning. This loop aligns with ReAct-style interleaving of reasoning and action \cite{yao2023react}, but adds a stronger separation between linguistic proposal and executable commitment. Such separation is essential in multi-agent settings because one agent's text may trigger another agent's action. The bridge between language and symbolic execution must therefore include validation, provenance, and state updates.

\subsection{Closest Verification-First Planning Work}

Formal-LLM supervises LLM plan generation with a developer-supplied automaton, directly addressing invalid plan sequences \cite{li2024formalllm}. Metagent-P combines LLM world knowledge, symbolic cognitive architecture, and metacognitive verification for open-world planning \cite{zhou2025metagentp}. Verified Multi-Agent Orchestration uses a plan--execute--verify--replan loop and an LLM-based completeness verifier over a question DAG \cite{zhang2026vmao}. The formal-methods position of Zhang et al. argues more broadly that trustworthy agents require integration between LLMs and formal methods \cite{zhang2025trustworthy}. These works make a generic ``neuro-symbolic'' novelty claim untenable; the remaining gap is narrower and operational.

Recent neighboring systems narrow that gap further. LOOP iteratively exchanges PDDL specifications and symbolic feedback rather than performing one-shot translation \cite{virwani2025loop}. Cornelio et al. combine hierarchical decomposition, knowledge-graph RAG, and a symbolic validator that also detects execution failure \cite{cornelio2025hierarchical}. Chimera combines an LLM strategist, symbolic constraints, and causal inference for multi-objective decision making \cite{akarlar2025chimera}. The distinction claimed here is therefore not neuro-symbolic planning in general. It is the joint treatment of heterogeneous agent admission, canonical authority, artifact contracts, refusal before dispatch, receiver-observed operation fencing, and consensus replication of the authorization state.

\begin{table}[htbp]
\centering
\caption{Closest verification-first work and the remaining thesis distinction.}
\label{tab:closest-verification-work}
\small
\begin{tabular}{p{0.20\textwidth}p{0.30\textwidth}p{0.38\textwidth}}
\toprule
\textbf{Work} & \textbf{Primary verification mechanism} & \textbf{Distinction of this thesis} \\
\midrule
Formal-LLM & Automaton-constrained plan generation & Authorizes composition against a live admitted registry, local authority, artifact contracts, and durable attempt evidence. \\
Metagent-P & Cognitive-symbolic planning plus metacognitive verification & Focuses on heterogeneous executor admission, refusal, protocol adaptation, and dependency-safe execution rather than one agent's open-world planning loop. \\
Verified Multi-Agent Orchestration & LLM verification of completeness followed by replanning & Uses deterministic plan-level gates for modeled predicates and distinguishes contract validity from LLM quality judgment. \\
Trustworthy-agent formal-methods position & Research agenda joining LLMs and formal methods & Supplies one concrete implementation refinement, crash-recovery boundary, and reproducibility protocol; it does not claim comprehensive formal verification. \\
\bottomrule
\end{tabular}
\end{table}

The bridge is not a single translator. It is better understood as a set of transformations and checks. Natural-language goals may be transformed into symbolic goals; symbolic plans may be transformed into natural-language explanations; tool schemas may be transformed into LLM-callable functions; monitor violations may be transformed into feedback for replanning. Each transformation can introduce error. Consequently, a robust bridge should record provenance, expose confidence or uncertainty when available, and support rejection or repair. This position differs from treating the LLM as a universal planner. It treats the LLM as one participant in a heterogeneous reasoning pipeline.

Hybrid reasoning also raises a granularity question. The LLM may operate at the level of high-level tasks, while a symbolic planner operates at the level of primitive actions. If the boundary is too high, the symbolic layer cannot catch relevant failures. If the boundary is too low, the LLM may be forced to produce brittle formal details it is not well suited to generate. A practical architecture should allow multiple levels of representation: strategic goals, decomposed tasks, protocol messages, symbolic constraints, and executable tool calls. Coordination then becomes the process of maintaining consistency across these levels.

\section{LLMs as Agent Components}

The rise of transformer-based language models created new possibilities for agent design. The transformer architecture enabled scalable sequence modeling through attention mechanisms \cite{vaswani2017attention}. Large pre-trained models demonstrated few-shot and instruction-following behavior \cite{brown2020language,ouyang2022training}. Chain-of-thought prompting showed that intermediate reasoning traces can improve performance on some reasoning tasks \cite{wei2022chain}. These developments made it practical to use LLMs as flexible reasoning and communication modules inside larger systems.

ReAct is an influential pattern because it interleaves reasoning traces with actions over external tools or environments \cite{yao2023react}. Toolformer and related work further demonstrate that language models can be adapted to call external tools \cite{schick2023toolformer}. Augmented language model surveys describe this broader movement toward coupling LLMs with retrieval, tools, memory, and external computation \cite{mialon2023augmented}. For autonomous agents, tool use changes the risk profile. A generated sentence may be harmless; a generated tool call can modify files, execute code, send messages, or change a physical state. Therefore, the agent architecture must treat tool invocation as an execution event governed by permissions, validation, and monitoring.

Memory is another key capability. Retrieval-augmented generation provides access to external knowledge \cite{lewis2020retrieval}; Reflexion introduces verbal feedback and episodic memory to improve future behavior \cite{shinn2023reflexion}; Voyager stores reusable skills learned in a Minecraft environment \cite{wang2023voyager}. These approaches show that agent performance depends not only on the base model but also on persistent context and learned artifacts. In a unified architecture, memory should be divided into forms with different semantics: conversational memory, factual knowledge, symbolic state, plan history, tool traces, and commitments. Collapsing all memory into unstructured text increases flexibility but reduces verifiability.

Embodied and interactive agents further illustrate both the promise and limits of LLM-based agency. Work on language models as zero-shot planners and embodied reasoning explores how LLMs can produce action sequences for simulated or physical tasks \cite{huang2022language,huang2022inner}. Voyager demonstrates open-ended exploration with a skill library \cite{wang2023voyager}. Generative Agents simulate social behavior through memory, reflection, and planning \cite{park2023generative}. These systems show that LLMs can support long-horizon behavior, but they also depend on scaffolding: memory retrieval, environment feedback, tool execution, and action validation. The agent is the full scaffolded system, not only the model.

The current LLM-agent literature also reveals a tension between generality and reliability. Generality comes from representing many tasks in language and relying on the model's broad prior knowledge. Reliability usually comes from narrowing the action space, adding tests, constraining tools, and validating intermediate artifacts. Autonomous systems need both. A useful agent may have to interpret open-ended human instructions, but it should not execute open-ended actions without controls. This tension explains why tool schemas, sandboxes, structured outputs, retrieval systems, and evaluators have become common in agent frameworks. They are attempts to recover reliability around a flexible linguistic core.

For multi-agent coordination, another limitation is social reasoning. LLMs can simulate dialogue and infer intentions, but simulated agreement does not necessarily correspond to coordinated execution. A planning agent may believe a worker agent has accepted a task because the worker responded positively, while the worker may lack the capability or authority to execute it. Classical MAS concepts such as capabilities, commitments, and organizational roles can reduce this ambiguity. The LLM can make role-based interaction more natural, but the system should still maintain explicit records of who can do what, who has promised what, and what evidence resolves the promise.

\section{LLM-Based Multi-Agent Systems}

LLM-based multi-agent systems adapt MAS ideas to teams of language-model-driven components. Surveys of LLM agents and LLM-based MAS identify common patterns: role specialization, debate, delegation, reflection, tool use, and conversation-based collaboration \cite{wang2024survey,guo2024llmmas,tran2025multiagent}. These systems often assign roles such as planner, critic, coder, reviewer, researcher, or executor. The roles are usually encoded in prompts rather than in formal organizational models, but the functional resemblance to classical MAS is clear.

AutoGen introduced a flexible framework for composing conversable agents that may use LLMs, tools, or humans in the loop \cite{wu2023autogen}. CAMEL uses role-playing prompts to study communicative cooperation among agents \cite{li2023camel}. MetaGPT encodes software-engineering workflows as standard operating procedures distributed across specialized agents \cite{hong2023metagpt}. ChatDev similarly frames software development as communication among role-specialized agents \cite{qian2023communicative}. Current engineering frameworks broaden this pattern: LangGraph provides low-level orchestration for durable, stateful agents \cite{langgraph2026}; Semantic Kernel provides middleware for integrating AI agents, models, plugins, and application code \cite{semantickernel2026}; CrewAI organizes autonomous agents into crews and flows for complex workflows \cite{crewai2026}. These systems demonstrate that multi-agent decomposition can improve task structure, distribute responsibilities, support persistence, and enable iterative critique. However, they do not by themselves settle whether a natural-language plan is authorized by an admitted registry, whether artifact contracts are semantically sufficient, or whether a refusal should be returned before dispatch.

Lightweight conversational-modeling frameworks address a narrower but important part of this landscape. Lingo is a type-safe Python framework for building LLM-powered applications around explicit conversational state, structured generation, structured flows, finite-state control, reflexive guardrails, and developer-controlled tool loops \cite{lingo2026}. Its relevance is not that it replaces a MAS platform, an interoperability protocol, or a symbolic planner. Rather, it shows a state-of-the-art pattern for disciplining the linguistic layer itself: free-form language is wrapped in typed context, constrained model decisions, schema-validated outputs, and state-conditioned dialogue. This makes Lingo a useful example of how linguistic agents can be engineered inside a larger autonomous system while still requiring external authorization and feasibility validation before generated proposals become executable commitments.

\begin{table}[htbp]
\centering
\caption{Representative LLM-based multi-agent approaches.}
\label{tab:llm-mas}
\begin{tabular}{p{0.20\textwidth}p{0.33\textwidth}p{0.34\textwidth}}
\toprule
\textbf{Approach} & \textbf{Coordination pattern} & \textbf{Limitation motivating hybrid design} \\
\midrule
AutoGen & Programmable conversations among agents, tools, and humans. & Conversation flexibility does not by itself provide formal protocol semantics. \\
CAMEL & Role-playing agents cooperate through prompted dialogue. & Role behavior is prompt-dependent and difficult to verify. \\
MetaGPT & Specialized agents exchange structured artifacts within a workflow. & Standard operating procedures require domain design and may remain task-specific. \\
ChatDev & Software development is decomposed into communicative roles. & Success depends on artifact consistency and evaluation of generated code. \\
LangGraph, Semantic Kernel, CrewAI & Practical orchestration, middleware, crews, flows, tools, and durable state for LLM applications. & Provide useful engineering substrates, but symbolic feasibility, authorization ownership, and infeasibility semantics remain application-level responsibilities. \\
Lingo & Typed conversational modeling with context, structured generation, flows, state machines, and reflexive guardrails. & Provides a disciplined linguistic layer, but not distributed MAS protocol semantics or symbolic feasibility checks. \\
\bottomrule
\end{tabular}
\end{table}

This framework landscape clarifies the difference between a linguistic-agent construction toolkit and a coordination architecture. The former can narrow model behavior into typed acts, support state-conditioned dialogue, and expose guardrail hooks; the latter must still decide which agents may act, whether a plan is feasible, how artifacts satisfy contracts, and what communication trace establishes completion. For this thesis, the SOTA implication is that language-mediated flexibility should be retained, but it should be placed behind explicit contracts, protocol state, and symbolic authorization.

The main lesson from this literature is that LLM-MAS research has reintroduced coordination as a practical necessity. When multiple LLM agents interact, the system needs role assignment, message routing, shared memory, conflict resolution, and evaluation of partial results. These are classical MAS problems. What is new is the linguistic richness of the agents and their ability to transform vague human instructions into candidate artifacts. The challenge is to connect that richness with formal models of interaction, state, and execution. Without such grounding, multi-agent dialogue may produce impressive transcripts but weak guarantees.

A second lesson is that role prompting is not the same as role semantics. In many LLM-MAS systems, an agent is assigned a role through natural-language instructions such as "act as a software architect" or "act as a reviewer." This can be effective for shaping model behavior, but it does not define a role in the organizational sense. A formal role has permissions, responsibilities, expected messages, authority boundaries, and accountability conditions. The gap between prompted roles and formal roles is significant for autonomous systems. A reviewer prompt may encourage criticism, but a reviewer role in a verified workflow should specify what evidence must be checked before approval.

The literature also suggests that multi-agent decomposition can sometimes improve output quality while also increasing system complexity. More agents mean more messages, more state transitions, more opportunities for inconsistency, and higher computational cost. The benefit of additional agents should therefore be justified by functional separation: one agent may specialize in natural-language interpretation, another in symbolic planning, another in verification, and another in execution. This division aligns with heterogeneous MAS more than with homogeneous debate among multiple copies of the same model. The thesis problem is thus not merely how to make many LLMs talk, but how to coordinate different kinds of reasoning entities.

\section{Modern Agent Interoperability Protocols}

The recent growth of agent frameworks has produced a new interoperability problem. Agents built in different frameworks may need to access tools, exchange context, advertise capabilities, delegate tasks, or collaborate across organizational boundaries. Several emerging protocols address parts of this problem. As of July 2026, the Model Context Protocol (MCP) defines a standardized client-server boundary for tools, resources, prompts, roots, sampling, elicitation, transports, and authorization; the thesis treats the 2025-11-25 specification as the stable cited baseline and the current draft page as protocol-evolution evidence rather than as a settled production guarantee \cite{anthropic2024mcp,modelcontextprotocol2025spec,modelcontextprotocol2026draft}. Agent2Agent (A2A) focuses on communication and interoperability between opaque agentic applications, including Agent Card discovery, JSON-RPC over HTTP(S), streaming, push notification, and structured data exchange; it is an open-source project under Linux Foundation governance, contributed by Google \cite{google2025a2a,a2aproject2025,linuxfoundation2025a2a}. The Agent Communication Protocol (ACP) was introduced as an HTTP-native agent communication standard, but its documentation and LF AI and Data describe ACP as part of A2A, with independent active development winding down \cite{ibm2025acp,acpdocs2026,lfaidata2025acp}. The Agent Network Protocol (ANP) emphasizes decentralized identity, discovery, messaging, and layered agent-network interoperability \cite{anp2025,chang2025anp}.

\begin{longtable}{p{0.16\textwidth}p{0.30\textwidth}p{0.38\textwidth}}
\caption{Emerging interoperability protocols for agent systems.}
\label{tab:protocols}\\
\toprule
\textbf{Protocol} & \textbf{Primary focus} & \textbf{Relevance to unified coordination} \\
\midrule
\endfirsthead
\toprule
\textbf{Protocol} & \textbf{Primary focus} & \textbf{Relevance to unified coordination} \\
\midrule
\endhead
MCP & Tool and context integration between AI applications and external systems. & Provides a standardized tool/context boundary, but does not by itself define full agent-to-agent social semantics. \\
A2A & Agent-to-agent communication, capability discovery, and collaboration. & Addresses cross-framework agent interaction and can support higher-level coordination protocols. \\
ACP & Lightweight HTTP-native agent communication, now aligned with A2A efforts. & Illustrates convergence toward shared messaging standards for heterogeneous agents. \\
ANP & Decentralized identity, discovery, and networked agent communication. & Highlights identity and discovery as first-class concerns in open agent ecosystems. \\
\bottomrule
\end{longtable}

These protocols should be interpreted carefully. They are important engineering developments, but they are not complete replacements for classical ACL semantics \cite{ehtesham2025survey}. MCP can expose tools to an LLM application, but it does not determine whether a tool call is a request, a delegation, a commitment, or a verified action. A2A can support agent discovery and communication, but application-level coordination still requires domain protocols and state models. ACP's convergence into A2A also shows that the protocol landscape is consolidating, not that the semantic coordination problem has disappeared. The relevant research question is therefore how modern protocol infrastructure can be mapped onto classical MAS concepts such as performatives, commitments, roles, and interaction protocols. A thesis architecture that bridges LLMs and symbolic AI should use modern protocols for interoperability while preserving explicit semantics for coordination and execution.

The protocol landscape also illustrates the difference between connectivity and coordination. Connectivity means that components can exchange messages or expose tools. Coordination means that those exchanges contribute to a coherent joint process with interpretable state and controlled effects. MCP improves the tool boundary by making external context and tools available through a common interface. A2A and related efforts address cross-agent messaging and capability discovery. These are necessary infrastructure layers, but a MAS still requires policies for when to delegate, how to evaluate capability claims, how to recover from failure, and how to resolve conflicting commitments. In this sense, modern protocols lower the cost of integration but do not remove the need for agent theory.

Protocol evolution is also a governance issue. If agents are expected to operate across organizations, they must advertise identity, capabilities, permissions, and trust assumptions. Decentralized discovery may support open ecosystems, but it also creates risks around impersonation, malicious agents, and incompatible policy regimes. A unified architecture should therefore separate protocol transport from trust decisions. It should be possible to discover an agent without automatically trusting it, to receive a request without accepting it, and to expose a tool without granting unrestricted authority. These distinctions are familiar in distributed systems and security engineering, but they must be made explicit in LLM-agent architectures.

\section{Safety, Verification, and Governance}

Safety and verification are central because autonomous agents can produce effects beyond text generation. Classical MAS and autonomous-systems research has long examined model checking, runtime verification, ethical decision constraints, and verification of agent programs \cite{fisher2013verifying,bordini2006model,dennis2016practical}. These techniques become more important when LLMs participate in action selection. A language model may produce a plausible action that violates a precondition, exceeds authority, leaks information, or conflicts with another agent's commitment. Symbolic verification cannot eliminate all risks, but it can check well-specified properties of plans, tool calls, and execution traces.

Runtime monitoring is a practical complement to offline verification. A monitor can enforce invariants over tool calls, memory writes, inter-agent messages, and environment actions. For example, a monitor may prevent an agent from executing a destructive operation without authorization, writing to protected memory, or sending a message that violates a protocol state. Such mechanisms fit naturally with a hybrid architecture: LLMs propose and explain, symbolic controllers validate and execute, and monitors observe the boundary between proposal and action.

LLM-specific safety risks include hallucination, prompt injection, tool misuse, data leakage, and manipulation of shared memory. Indirect prompt injection shows that external content can influence tool-using LLM applications in ways not intended by the user or developer \cite{greshake2023not}. Security analyses of MCP ecosystems emphasize that tool integration expands the attack surface for AI applications \cite{hou2025mcpsecurity}. In multi-agent settings, these risks can propagate: one compromised agent may poison shared memory, issue misleading messages, or delegate unsafe actions. Therefore, governance must include identity, permissions, audit logs, provenance, and explicit boundaries between agents.

The safety literature supports a defense-in-depth position. Natural-language policies are useful for communication, but enforceable policies require structured representations. Human oversight is important, but scalable systems also need automated checks. LLM self-critique can identify some errors, but independent symbolic validation is more reliable for formal constraints. For this reason, the state of the art points toward architectures where language models are powerful but bounded participants in a larger verified control system.

Verification must also be scoped realistically. Full verification of an open-ended LLM is not currently practical. However, many safety-relevant properties concern the symbolic envelope around the model rather than the model weights themselves. A system can verify that a tool call conforms to a schema, that a plan action satisfies a precondition, that a workflow state permits a transition, that a user has authorized a capability, or that a message follows a protocol. These checks do not prove that the LLM is truthful or safe in all contexts, but they reduce the space in which model errors can produce harmful effects. The thesis can therefore pursue practical assurance by verifying the controllable parts of the agent loop.

Governance also requires traceability. A coordinated system should record why an action was selected, which agent proposed it, which module validated it, which policy authorized it, and what result followed. Such traces support debugging, evaluation, and accountability. They also support scientific analysis: without traces, it is difficult to determine whether success came from genuine coordination, accidental model knowledge, or hidden human intervention. For a master's thesis solution, traceability is valuable both as an engineering feature and as an empirical research instrument.

\section{Evaluation and Benchmarks}

Evaluation of unified LLM-symbolic multi-agent systems requires multiple dimensions. Classical MAS evaluation considers coordination efficiency, communication overhead, task allocation quality, robustness to agent failure, scalability, and social welfare. LLM-agent evaluation considers task success, reasoning quality, tool-use accuracy, factuality, cost, latency, robustness, and safety. A unified system must be evaluated across both sets of criteria. A system that produces elegant dialogue but fails execution is inadequate; a system that executes reliably but cannot interpret natural-language goals may also fail the intended use case.

Several benchmarks have shaped LLM-agent evaluation. AgentBench evaluates LLMs as agents across interactive tasks \cite{liu2023agentbench}. WebArena tests web navigation and task completion in realistic web environments \cite{zhou2023webarena}. ALFWorld aligns text instructions with embodied environments \cite{shridhar2021alfworld}. SWE-bench evaluates the ability to resolve real-world software issues \cite{jimenez2024swebench}. GAIA emphasizes general assistant tasks requiring reasoning and tool use \cite{mialon2023gaia}. OSWorld evaluates computer-control agents in operating-system environments \cite{xie2024osworld}. These benchmarks are useful, but they do not fully capture the problem of heterogeneous multi-agent coordination with symbolic verification.

\begin{table}[htbp]
\centering
\caption{Evaluation dimensions for unified LLM-symbolic multi-agent systems.}
\label{tab:evaluation}
\begin{tabular}{p{0.25\textwidth}p{0.65\textwidth}}
\toprule
\textbf{Dimension} & \textbf{Representative questions} \\
\midrule
Task performance & Does the system achieve the specified goal, and with what cost, latency, and solution quality? \\
Coordination quality & Are tasks allocated effectively, messages grounded, conflicts resolved, and commitments tracked? \\
Symbolic validity & Do generated plans satisfy preconditions, constraints, and domain rules before execution? \\
Robustness & Does performance persist under rephrased goals, partial observability, agent failure, or adversarial inputs? \\
Safety and governance & Are tool calls authorized, logged, monitored, and constrained by explicit policies? \\
\bottomrule
\end{tabular}
\end{table}

The evaluation gap is therefore clear. Existing benchmarks evaluate individual LLM agents, web navigation, embodied tasks, coding tasks, or general assistant behavior. Classical MAS benchmarks evaluate coordination mechanisms under formal assumptions. Fewer evaluations measure whether LLM-mediated agents can coordinate through natural language while maintaining symbolic state, formal commitments, and verified execution. This gap motivates thesis-specific scenarios that test not only whether a final task is completed, but also whether the coordination trace is interpretable, valid, and safe.

A useful evaluation design should therefore include both outcome metrics and process metrics. Outcome metrics measure whether the task was completed, how long it took, how much it cost, and whether the output met quality criteria. Process metrics measure whether the correct agent accepted the task, whether the symbolic plan was valid, whether messages followed protocol, whether tool calls were authorized, and whether failures were detected at the proper boundary. This distinction is important because an LLM system can sometimes achieve a correct result through an unsafe or non-generalizable process. Scientific evaluation should avoid rewarding such accidental success without qualification.

Evaluation should also include ablations. A unified architecture can be compared against variants that remove the symbolic planner, remove the LLM interpreter, remove protocol tracking, or remove runtime monitoring. Such comparisons make it possible to test whether each component contributes to performance, safety, or interpretability. Without ablation, the architecture may appear effective only because the task is easy or because a single model performs most of the work. The state of the art in agent evaluation increasingly recognizes this issue, but unified LLM-symbolic MAS evaluation remains underdeveloped.

\section{Synthesis and Research Gap}

The reviewed literature shows that the components needed for unified coordination already exist, but they remain insufficiently integrated. Classical agent theory defines autonomy, interaction, commitment, planning, and verification. MAS research provides mechanisms for task allocation, negotiation, shared-state coordination, norms, and distributed constraint reasoning. Symbolic AI supplies formal representations for knowledge, plans, rules, and control. LLM research supplies linguistic interpretation, flexible communication, tool-use scaffolding, and broad task generalization. Modern interoperability protocols supply practical interfaces for tools, context, and cross-agent communication.

\begingroup
\small
\setlength{\tabcolsep}{3pt}
\begin{longtable}{
>{\raggedright\arraybackslash}p{0.18\textwidth}
>{\raggedright\arraybackslash}p{0.23\textwidth}
>{\raggedright\arraybackslash}p{0.23\textwidth}
>{\raggedright\arraybackslash}p{0.24\textwidth}}
\caption{Synthesis matrix deriving requirements for the proposed framework.}
\label{tab:synthesis-matrix}\\
\toprule
\textbf{Research line} & \textbf{Contribution} & \textbf{Limitation} & \textbf{Derived thesis requirement} \\
\midrule
\endfirsthead
\toprule
\textbf{Research line} & \textbf{Contribution} & \textbf{Limitation} & \textbf{Derived thesis requirement} \\
\midrule
\endhead
Classical agents and MAS & Autonomy, roles, task allocation, negotiation, commitments, and coordination failures. & Does not by itself handle open-ended natural-language intent or modern LLM tool scaffolds. & Treat agents as accountable participants with explicit state, capabilities, and commitments. \\
ACLs and interaction protocols & Performatives, protocol state, content languages, and observable communicative force. & Structured acts are often absent from free-form LLM dialogue. & Map language-mediated exchanges to typed requests, acceptances, reports, and completion evidence. \\
Symbolic AI & Plans, constraints, schemas, rules, and verification conditions. & Only reasons over modeled facts, actions, and contracts. & Use symbols for authorization, compatibility, evidence, and refusal rather than for all semantic interpretation. \\
Neuro-symbolic and LLM planning & Translation between natural language, tools, formal goals, and external solvers. & LLM-generated plans can remain plausible but invalid under action semantics. & Separate candidate proposal from feasibility checking and executable commitment. \\
LLM-MAS frameworks & Role-based decomposition, critique, tool use, and flexible collaboration. & Prompted roles and conversation traces do not guarantee role semantics or accountability. & Require explicit ownership, authority, traceability, and process metrics beyond final output. \\
Modern interoperability protocols & Tool/context access, agent discovery, task exchange, and cross-framework connectivity. & Connectivity is not coordination, and discovery is not trust. & Keep protocol transport subordinate to coordination semantics, trust decisions, and state models. \\
Safety and evaluation & Monitoring, authorization, audit logs, benchmark practice, and process/outcome separation. & Full verification of open-ended LLM behavior remains impractical. & Verify the symbolic envelope: contracts, permissions, transitions, traces, and bounded auxiliary capabilities. \\
\bottomrule
\end{longtable}
\endgroup

The state of the art nevertheless leaves several unresolved issues. First, the term agent is used inconsistently across classical MAS and LLM literature, which obscures the distinction between a language model, a tool-using application, and an autonomous system. Second, LLM-based coordination often relies on free-form conversation without explicit performatives, commitments, or protocol state. Third, symbolic planning and verification are often treated as optional tools rather than as structural components of safe execution. Fourth, emerging protocols improve connectivity but do not, by themselves, solve the semantic problem of coordination. Finally, benchmark practice has not yet converged on evaluation methods for systems that combine linguistic agents, symbolic agents, tool agents, and autonomous execution components.

These gaps define the thesis positioning. A unified coordination framework should treat LLMs as linguistic reasoning modules embedded inside full autonomous agents. It should translate between natural-language interaction and symbolic coordination structures. It should distinguish proposals from commitments, and commitments from executed actions. It should expose tools through controlled interfaces, represent shared state explicitly, and validate plans before execution. Such a framework does not reject the flexibility of LLMs; rather, it makes that flexibility usable in autonomous systems where coordination, accountability, and safety matter.

The resulting research direction is not a competition between neural and symbolic AI. It is an architectural question about how heterogeneous reasoning components should cooperate. LLMs are valuable because they make autonomous systems more communicative, adaptive, and accessible to human intent. Symbolic AI is valuable because it makes plans, constraints, commitments, and verification conditions explicit. MAS research is valuable because it explains how autonomous entities interact when no single component has all information or authority. The contribution of a unified framework is to place these elements in a coherent coordination loop, where language supports interpretation, symbols support execution semantics, and protocols support collective behavior.
