\chapter{Conclusions and Future Work}
\label{chap:conclusion}

The thesis proposes a protocol-adaptable boundary between linguistic semantic selection and executable multi-agent coordination. RQ1 is answered at the design and prototype level: canonical capabilities, typed constraints, task dependencies, authority, validation and completion contracts, artifacts, attempts, and persisted trace identity must be explicit to justify dispatch. RQ2 is answered architecturally: language selects admitted goal, policy, contract, exclusion, and ambiguity identifiers, while deterministic components own dependency closure, typed-request construction, provider assignment, enforceable predicates, authorization, refusal, and recording. RQ3 is answered through bounded task-local auxiliary specifications whose method, schema, validator, read-only authority, and lifecycle are explicit.

RQ4 is answered by the complete frozen v0.7 matrix. At primary seed 11, the hybrid achieves 71.88\% balanced accuracy, 44.79\% feasible recall, and one unsafe acceptance; the direct LLM achieves 53.13\% balanced accuracy, 98.96\% feasible recall, and 89 unsafe acceptances. Safety superiority is supported by the paired interval, but zero unsafe acceptance, useful coordination, recall non-inferiority, and lexical superiority all fail. The exact-name/alias baseline solves all 64 designed cases, so the corpus does not establish a need for an LLM and exposes a strong construct-validity limitation. Only two of six predeclared gates pass: safety superiority and the deterministic symbolic invariants. No independent adjudication or broad superiority is claimed.

RQ5 is partially supported by the complete consensus-v4 campaign. The implementation supplies consensus-backed storage, generation-CAS configuration, serialized resumable membership operations, fencing, linearizable authorization reads, receiver-side operation controls, and fail-closed quorum responses. Forty-three of 45 trials passed and no executed primary safety invariant failed; 140 of 150 primary checks executed. However, one repetition of each reconfiguration path ended in an infrastructure timeout before its checks ran. Moreover, the preserved dirty worktree makes \texttt{evidence\_valid=false}, so the campaign is descriptive complete evidence rather than accepted clean release evidence. No claim is made for multi-host availability, Byzantine tolerance, automatic recovery after majority loss, or exactly-once external effects.

The principal scientific claim is therefore deliberately narrow. The implementation demonstrates how established symbolic, security, persistence, interoperability, and crash-consensus mechanisms can be integrated so that unsupported linguistic proposals are refused before action and authoritative mutations stop when quorum is unavailable. Duplicate external effects are suppressed only when receivers enforce both operation keys and monotonic fences. The work does not prove semantic truth, production safety, malicious-agent resistance, decentralized authorization, Byzantine tolerance, exactly-once effects, or universal protocol interoperability.

The technical evidence includes unsuccessful outcomes rather than selected reruns. It does not establish that an LLM is necessary: the exact-alias baseline solves the full v0.7 corpus, while Gemma's production hybrid refuses every feasible primary-seed case. The deterministic compiler nevertheless passes all oracle, registry-permutation, alternative-provider, and bounded-search invariants, and eight observed hybrid acceptances complete fixture A2A replay. Remaining release work is administrative and provenance-related: supply tutor and tribunal metadata, create a clean packaging commit after author review, rerun consensus-v4 from that clean source, rebuild the signed-off PDF, tag, push, and archive the submission package. Longer-term research should independently source and adjudicate a broader raw-language corpus with low lexical overlap, improve useful semantic admission without weakening fail-closed behavior, evaluate larger provider-search spaces, investigate the two reconfiguration timeouts, introduce mTLS with per-node identities and rotation, test external domain agents, add semantic validators and replanning, place voters in independent failure domains, and evaluate snapshot and disaster-recovery procedures in broader deployments.
