\chapter{System Design: Coordination Agent over SDK Middleware}
\label{chap:implementation-design}

This chapter describes how the protocol-independent framework is realized as an executable Python prototype. Its goal is to preserve the theoretical separation between non-authoritative linguistic proposal, symbolic authorization, middleware-mediated dispatch, and durable evidence while making every important abstraction traceable to a concrete component, record, state transition, or deployment boundary.

The chapter begins with the framework instantiation and component responsibilities. It then covers agent admission and communication, the minimum data structures, coordinator workflow, and local and linguistic specializations. The architecture and execution-sequence sections show how requests cross the SDK boundary, after which failure handling and consensus-backed coordinator nodes explain runtime and distributed behavior. A final synthesis relates the implementation back to the theory and states the remaining operational limits.

\section{Instantiation of the Framework}

The protocol-independent theory from Chapter~\ref{chap:theoretical-solution} is instantiated as a Python coordination agent built over a middleware SDK. The theoretical contribution does not require a particular transport; for the concrete prototype, the implementation selects the Agent2Agent (A2A) protocol as one interoperability adapter. A2A is an open protocol intended to support communication and coordination among agents built with different technologies. Its project documentation describes standardized communication, Agent Cards for discovery, flexible interaction patterns, and rich data exchange through text, files, and structured JSON \cite{a2aproject2025}. Google's announcement similarly frames A2A as an open protocol for secure information exchange and coordination among agents across enterprise platforms \cite{google2025a2a}.

The implementation separates two roles. \texttt{CoordinationSdk} is the communication and runtime middleware. It registers remote A2A agents, local Python agents, and Lingo-backed linguistic agents; normalizes their capabilities; dispatches authorized tasks; converts artifacts; records runtime failures; and exposes communication traces. It does not plan, decide feasibility, authorize decompositions, or coordinate complete user requests.

\texttt{CoordinationAgent} is the central coordinating role. It receives the user request, uses the SDK to inspect the registry and capability index, invokes SDK-registered linguistic capabilities when interpretation or candidate planning is needed, builds direct or decomposed candidate plans, calls the symbolic \texttt{FeasibilityAnalyzer}, authorizes or refuses the plan, dispatches only authorized tasks through the SDK, and aggregates the final result. The agent never calls raw A2A transports, raw local handlers, or raw Lingo runtimes directly. All execution traffic still passes through \texttt{CoordinationSdk}.

A2A remains the default wire model for independently deployed agents. Remote agents are admitted through A2A Agent Cards and invoked through A2A-compatible task exchange. Local Python agents and Lingo-based linguistic agents are registered through the SDK and normalized into the same internal agent records, with identity, capabilities, input and output modes, trust metadata, invocation endpoint, and validation contract. Local and linguistic agent authors are therefore not required to implement A2A plumbing, while the coordination agent can reason over a uniform registry.

The prototype also uses Lingo as an optional linguistic and context-engineering substrate \cite{lingo2026}. Lingo is not hidden inside the SDK as a coordinator. Instead, Lingo-backed interpretation, extraction, classification, or candidate-planning behavior is exposed as SDK-registered linguistic capabilities. Non-linguistic local agents can use the SDK without depending on Lingo. The prototype dependency set pins \texttt{lingo-ai==2.0.2} and uses \texttt{a2a-sdk}, \texttt{pydantic}, \texttt{httpx}, \texttt{FastAPI}, \texttt{Uvicorn}, \texttt{pytest}, and \texttt{pytest-asyncio} for the executable implementation, service wrapper, and tests.

The version 0.4 bridge narrows candidate planning further. The model returns a \texttt{LinguisticPlanDraft} containing only canonical requirement/capability selections, dependencies expressed between requirement identifiers, unresolved terms, and rationale. It cannot generate task identifiers, providers, trust, credentials, contracts, schemas, artifacts, or authority. A deterministic \texttt{PlanHydrator} verifies complete and unique requirement coverage, exact capability identity, provider availability, dependency references, and acyclicity; it then selects the lexicographically first admitted provider and constructs tasks, assignments, validation contracts, artifact obligations, execution order, and completion contract from the admitted request and registry. Hydration errors may support one bounded repair attempt, after which failure remains closed. Direct deterministic planning still precedes this linguistic fallback when exact matching already decides the plan.

Three runtime mechanisms are also explicit secure-default policies: dependency dispatch requires successful predecessors, auxiliary admission requires eligibility, plan-local lifecycle, non-persistence, validation, and read-only authority, and durable trace evidence is exposed from persisted store events. Unsafe alternatives exist only as named experimental negative controls in the runtime-ablation runner; production constructors select the strict policies by default. The trace negative control removes returned diagnostic evidence without deleting state-critical events, leases, fencing, attempts, or terminal records.

For deployment and reproducibility, the repository includes a thin HTTP service and Docker configurations around the same \texttt{CoordinationAgent} and \texttt{CoordinationSdk} objects. The service exposes planning and coordination endpoints, separate liveness and readiness probes, cluster status, metrics, and authenticated internal membership and registry operations. It does not bypass the SDK boundary or change the theoretical ownership of planning and authorization. \texttt{COORDINATION\_STORE\_URL} selects JSONL for local use, PostgreSQL for legacy shared-store compatibility, or etcd for consensus-backed authoritative state. When etcd is authoritative, PostgreSQL is optional best-effort asynchronous audit storage and its failure does not make the coordinator unready.

The mapping between formal objects and implementation records is direct. The records defined in Section~\ref{sec:data-structures} materialize the formal objects of Chapter~\ref{chap:theoretical-solution} as follows.

\begin{itemize}
  \item \texttt{CoordinationAgent} materializes the central coordinating authority. It owns interpretation workflow, direct matching, decomposition, feasibility checks, auxiliary decisions, authorization, and final result construction.
  \item \texttt{CoordinationSdk} materializes the communication abstraction. It owns registry admission, A2A gateway behavior, local and Lingo runtimes, task dispatch, artifact normalization, runtime failure recording, and communication trace access.
  \item \texttt{CoordinationStore} materializes durable coordination state. Its JSONL and PostgreSQL adapters preserve local and legacy backends. Its etcd adapter is authoritative in the fault-tolerant topology and stores sessions, events, leases, registry snapshots, plan generations, commitments, attempts, artifacts, and terminal results through revision-checked transactions.
  \item \texttt{AgentRegistryEntry} materializes an agent $a$ (Definition~\ref{def:agent}), supplying $\mathit{id}$, declared capabilities $C(a)$, input modes $M_{\mathrm{in}}(a)$, output modes $M_{\mathrm{out}}(a)$, local admission policy $\tau(a)$, observable status $\sigma(a)$, and the invocation endpoint used by the SDK.
  \item \texttt{CapabilityRequirement} materializes a required capability descriptor $q \in \mathcal{Q}$ with modalities, schemas, constraints, trust level, and auxiliary eligibility.
  \item \texttt{ConstraintSpec} materializes one deterministic predicate with a stable identifier, declared evidence source, JSON-pointer path, typed operator, expected value, and optional requirement/task scope. Legacy prose is readable but unresolved and therefore cannot authorize.
  \item \texttt{ValidationContract} combines complete JSON Schema validation, required artifacts, kinds, fields, evidence types, and an optional registered validator. \texttt{CompletionContract} defines required terminal task states, final artifacts, and validator success.
  \item \texttt{ProblemRequest} materializes the request $P = (g, \mathcal{Q}, \chi, \mathcal{O}, \Gamma)$ from Definition~\ref{def:problem-request}.
  \item \texttt{SolutionProposal} materializes the coordination plan $\Pi$ from Definition~\ref{def:coordination-plan}, including task requirements, assignments, dependencies, auxiliary specifications, expected artifacts, and completion criteria.
  \item \texttt{FeasibilityReport} records the verdict and predicate evidence from Definition~\ref{def:feasibility}. It is produced by the \texttt{FeasibilityAnalyzer} at the \texttt{CoordinationAgent} layer.
  \item \texttt{GeneratedNlpAgentSpec} materializes the auxiliary capability specification from Definition~\ref{def:auxiliary}, bounded by Rules~\ref{rule:approved-construction}--\ref{rule:lifecycle}.
  \item \texttt{CoordinatorState} is the coordination agent's session state containing the registry snapshot, interpreted request, candidate plan, feasibility evidence, and trace.
  \item The recovery records make durable state explicit: \texttt{SessionState}, \texttt{TaskState}, \texttt{PlanGeneration}, and \texttt{TaskCommitment} describe coordination progress, while \texttt{TaskAttemptRecord}, \texttt{LeaseRecord}, and \texttt{TerminalResultRecord} support invariant checks.
  \item \texttt{TraceEvent} records auditable observations across interpretation, authorization, delegation, and artifact handling.
\end{itemize}

Each of the five properties stated in Chapter~\ref{chap:theoretical-solution} (justified acceptance, simplicity, bounded auxiliary synthesis, traceability, and explicit refusal) corresponds to a concrete decision or trace event owned by the \texttt{CoordinationAgent}, supported by communication evidence supplied through the SDK.

\section{System Components}

The proposed implementation is organized around a coordinating agent above a middleware SDK. The \texttt{CoordinationAgent} is the planner and authorizer. It receives user requests, maintains coordination state, requests linguistic interpretation when needed, constructs candidate plans, applies deterministic feasibility checks, decides whether to authorize or refuse the plan, dispatches authorized work through the SDK, and aggregates the final result.

\texttt{CoordinationSdk} is the communication and runtime substrate. It is the facade through which agents are registered and invoked. The SDK may use private helper classes internally, but adapter inheritance and transport details are not part of the public API. The SDK can reject invalid configuration, malformed registrations, missing handlers, invalid Agent Cards, and transport-level failures, but it does not decide whether a user request is feasible as a multi-agent plan.

The private SDK subsystems keep communication responsibilities simple. The Registry and Capability Index maintain admitted agent records and searchable capability metadata. The A2A Gateway fetches and validates Agent Cards, sends A2A tasks, and normalizes remote status, artifacts, cancellations, timeouts, and failures. The Local Agent Runtime wraps Python functions, scripts, classes, symbolic planners, and deterministic tools. The Lingo Agent Runtime wraps Lingo apps, flows, tools, state machines, and guardrails when linguistic behavior is required. The SDK Trace Store records communication and runtime observations. The Artifact Normalizer converts outputs into validator-friendly records.

\begin{table}[htbp]
\centering
\caption{Implementation components mapped to theoretical concepts.}
\label{tab:implementation-components}
\begin{tabular}{p{0.30\textwidth}p{0.58\textwidth}}
\toprule
\textbf{Component} & \textbf{Role in the implementation} \\
\midrule
\texttt{CoordinationAgent} & Owns interpretation workflow, direct matching, decomposition, feasibility checks, auxiliary decisions, authorization, and final result construction. \\
\texttt{CoordinationSdk} & Exposes middleware operations for registration, registry inspection, capability indexing, task dispatch, artifact normalization, runtime failure reporting, traces, and session reset. \\
\texttt{FeasibilityAnalyzer} & Applies deterministic checks for the \texttt{CoordinationAgent}; it is not part of the SDK's communication boundary. \\
A2A Gateway & Handles Agent Card ingestion and A2A-compatible communication with remote agents. \\
Local Agent Runtime & Wraps Python handlers and deterministic local tools as SDK-managed agent capabilities. \\
Lingo Agent Runtime & Wraps optional Lingo apps, flows, tools, state machines, and guardrails as ordinary SDK capabilities. \\
SDK Trace Store & Tracks communication events, runtime failures, task status, and artifact observations. \\
Artifact Normalizer & Converts remote, local, and linguistic outputs into records usable by validators and aggregators. \\
Auxiliary Capability Factory & Used by the \texttt{CoordinationAgent} to specify temporary lightweight linguistic agents when justified. \\
Coordination Store & Persists session events, leases, fencing tokens, plan generations, task commitments, attempts, artifacts, terminal results, and recovery state. \\
\bottomrule
\end{tabular}
\end{table}

\section{Agent Registration and Communication Boundary}

The SDK admits three kinds of agents through one communication boundary. A remote A2A agent is registered from an Agent Card URL. \texttt{AgentAdmissionPolicy} checks allowed origin, HTTPS policy, locally assigned trust, required credential schemes, and the explicit development-only HTTP exception. Effective trust is never accepted from the remote card. A credential provider supplies ephemeral request headers, while reports retain only scheme names. The SDK preserves card security schemes, canonical skill identifiers, endpoints, and modes in the internal registry entry.

A local agent is registered with \texttt{register\_local\_agent(name, capabilities, handler, ...)}. The handler may be a Python function, script wrapper, class, symbolic planner, or deterministic tool. The local agent does not need to know about A2A. The SDK supplies the same identity, capability, modality, authority, invocation, and validation metadata that a remote Agent Card would otherwise provide. During dispatch, the SDK invokes the local handler through the Local Agent Runtime and records the result as a normal task observation.

A linguistic agent is registered through the SDK's linguistic-agent registration method. The wrapped object may be a Lingo app, flow, structured generation routine, state machine, tool set, or guardrail-bearing conversational process. The SDK exposes the linguistic agent as an ordinary capability provider. Lingo remains optional and internal to that capability. The \texttt{CoordinationAgent} may invoke the linguistic capability through \texttt{sdk.send\_task(...)} when it needs interpretation, extraction, classification, explanation, or candidate planning.

No registered agent communicates directly with another registered agent through application-level shortcuts. The \texttt{CoordinationAgent} also does not bypass the SDK to reach remote transports, local handlers, or Lingo runtimes. This preserves one communication boundary while keeping planning and feasibility outside the SDK.

\section{Minimum Data Structures}
\label{sec:data-structures}

The implementation can be described through two public surfaces and a set of typed conceptual records. The records are represented as Pydantic models in the prototype so that Lingo structured generation, deterministic validation, SDK registration, coordination-agent planning, and test fixtures use the same schema surface. The records are not tied to a deployment platform; they define the information that the prototype must preserve.

\begin{Verbatim}[fontsize=\small,breaklines=true,breakanywhere=true]
CoordinationSdk:
  register_a2a_agent(agent_card_url, trust_level="standard")
  register_local_agent(name, capabilities, handler, ...)
  register_linguistic_agent(name, capabilities, lingo_app_or_flow, ...)
  registry_snapshot()
  capability_index()
  send_task(agent_id, task)
  trace()
  reset_session()

CoordinationAgent:
  coordinate(user_request, context=None)
  check_feasibility(problem_or_plan)
  trace()

AgentRegistryEntry:
  agent_id
  name
  agent_kind
  service_endpoint
  invocation_endpoint
  description
  skills
  input_modes
  output_modes
  status
  trust_level
  validation_contract
  source_card

CapabilityRequirement:
  name
  requirement_id
  capability_id
  description
  side_effect_class
  input_schema
  output_schema
  input_modes
  output_modes
  constraints
  auxiliary_eligible
  required_trust_level
  validation_contract

ConstraintSpec:
  constraint_id
  source
  path
  operator
  expected
  requirement_id
  task_id

ValidationContract:
  json_schema
  required_artifacts
  artifact_kinds
  required_fields
  evidence_types
  validator_id

CompletionContract:
  required_task_states
  required_artifacts
  require_all_task_validators

ProblemRequest:
  user_goal
  requirements
  constraints
  required_artifacts
  context

TaskSpec:
  task_id
  requirement_name
  requirement_id
  capability_id
  assigned_to
  auxiliary_spec_id
  depends_on
  expected_artifacts
  validation_contract

FeasibilityReport:
  feasible
  matched_agents
  generated_nlp_agents
  missing_capabilities
  risks
  explanation
  evidence

SolutionProposal:
  tasks
  generated_nlp_agents
  selected_agents
  execution_order
  expected_artifacts
  completion_criteria
  completion_contract
  constraint_evidence

GeneratedNlpAgentSpec:
  purpose
  input_schema
  output_schema
  method
  validation_rule
  lifecycle
  authority_bounds
  persists

CoordinatorState:
  registry_snapshot
  interpreted_request
  candidate_plan
  feasibility_evidence
  trace

TraceEvent:
  event_type
  message
  session_id
  plan_id
  task_id
  attempt_id
  source
  data
  timestamp

SessionState:
  new | planning | authorized | dispatching
  recovering | completed | failed | infeasible

TaskState:
  pending | running | completed | failed
  timeout | skipped | unknown

LeaseRecord:
  session_id
  holder_id
  fencing_token
  expires_at
  heartbeat_at

TaskAttemptRecord:
  session_id
  plan_id
  task_id
  attempt_id
  idempotency_key
  coordinator_id
  fencing_token
  state

TaskExecutionResult:
  task_id
  agent_id
  agent_kind
  status
  output
  artifacts
  error
  metadata

CoordinationRunResult:
  status
  plan_result
  task_results
  artifacts
  explanation
  trace
\end{Verbatim}

The \texttt{CoordinationSdk} public surface is intentionally limited to registration, inspection, dispatch, trace access, and session control. It hides the differences among remote A2A agents, local Python agents, and Lingo-backed linguistic agents while preserving enough metadata for the coordination agent to apply feasibility rules uniformly. The \texttt{CoordinationAgent} public surface contains the complete coordination workflow and feasibility query. The \texttt{AgentRegistry\allowbreak Entry} materializes the discoverable agent set. The \texttt{Capability\allowbreak Requirement} and \texttt{Problem\allowbreak Request} represent the user's goal, constraints, required capabilities, and required artifacts. The \texttt{Solution\allowbreak Proposal} is a candidate plan, not an executable plan until accepted. The \texttt{Feasibility\allowbreak Report} explains whether the coordination agent can solve the problem and why. \texttt{TaskExecutionResult} and \texttt{CoordinationRunResult} separate runtime observations from the pre-execution feasibility decision. The \texttt{Generated\allowbreak Nlp\allowbreak Agent\allowbreak Spec} represents a temporary linguistic capability, such as a classifier or extractor, that can be created for the current solution without becoming a permanent system agent. The \texttt{Coordinator\allowbreak State} and \texttt{Trace\allowbreak Event} records provide the auditable ledger required by the traceability property.

\section{Coordinator Workflow}

The application-level coordinator is \texttt{CoordinationAgent}. It talks to the distributed system only through \texttt{CoordinationSdk}, but it keeps planning and authorization outside the SDK. This means the SDK is reusable for non-coordinating clients, while the thesis contribution remains concentrated in the coordination agent's feasibility-aware workflow.

\begin{Verbatim}[fontsize=\small,breaklines=true,breakanywhere=true]
sdk = CoordinationSdk(...)
analyzer = FeasibilityAnalyzer(...)
agent = CoordinationAgent(
    sdk=sdk,
    feasibility_analyzer=analyzer,
    linguistic_interpreter="optional-sdk-agent-id",
    auxiliary_factory=optional_factory,
    artifact_aggregator=aggregator,
)

sdk.register_a2a_agent(agent_card_url, trust_level="standard")
sdk.register_local_agent(name, capabilities, handler, ...)
sdk.register_linguistic_agent(name, capabilities, lingo_app_or_flow, ...)

result = agent.coordinate(user_request, context=session_context)
\end{Verbatim}

Inside \texttt{coordinate}, the \texttt{CoordinationAgent} performs the simple-first decision process. It asks the SDK for the current registry and capability index, transforms or admits the user's message into a typed \texttt{ProblemRequest}, attempts direct capability matching, requests candidate decomposition only when direct delegation is insufficient, checks the proposal with the \texttt{FeasibilityAnalyzer}, adds a bounded auxiliary specification only when the feasibility model permits it, and dispatches authorized tasks through the SDK.

\begin{Verbatim}[fontsize=\small,breaklines=true,breakanywhere=true]
CoordinationAgent.coordinate(problem):
  registry = sdk.registry_snapshot()
  capabilities = sdk.capability_index()
  request = interpret_or_admit_problem(problem, sdk)

  candidate_plan = direct_or_decomposed_plan(request, capabilities, sdk)
  report = feasibility_analyzer.check(request, registry, candidate_plan)

  if report.has_auxiliary_eligible_gap:
      auxiliary = auxiliary_factory.specify(report.gap)
      candidate_plan = candidate_plan.with_auxiliary_agent(auxiliary)
      report = feasibility_analyzer.check(request, registry, candidate_plan)

  if report.is_not_feasible:
      return structured_infeasibility(report)

  results = [
      sdk.send_task(task.assigned_to, task)
      for task in candidate_plan.execution_order
  ]
  return artifact_aggregator.combine(results)
\end{Verbatim}

Task dispatch remains polymorphic inside the SDK but not visible to the coordination agent. If the selected executor is a remote A2A agent, the A2A Gateway sends the task and normalizes the remote response. If it is a local agent, the Local Agent Runtime invokes the registered handler and converts the result into the same task observation shape. If it is a Lingo-backed agent, the Lingo Agent Runtime executes the approved app, flow, tool, or state machine and returns a normalized artifact. In all cases, planning infeasibility is structured data returned by the \texttt{CoordinationAgent}. SDK exceptions are reserved for invalid SDK configuration or genuine transport/runtime failures.

This workflow keeps the LLM useful but bounded. Lingo may supply typed context handling and structured proposal mechanisms through SDK-registered linguistic capabilities, but it does not alone decide feasibility. Deterministic checks remain responsible for explicit properties such as availability, modality compatibility, declared skills, authority, and missing capabilities. Lingo's native tool-calling interface may be useful for controlled experiments that intentionally remove symbolic checks, but the main hybrid implementation uses developer-controlled structured output so that symbolic authorization remains explicit.

\section{Local and Linguistic Agent Specializations}

The SDK treats local and linguistic agents as specializations of the same communication boundary rather than as separate architectural paths. A local agent is appropriate when the capability is deterministic, symbolic, scriptable, or already implemented in Python. Examples include a rule-based planner, a schema validator, a database wrapper, a file transformer, or a mathematical tool. The SDK supplies the communication wrapper, so the local agent author focuses on the capability contract and handler behavior.

A linguistic agent is appropriate when the capability transforms, interprets, classifies, summarizes, or produces language-mediated artifacts. It may be implemented with Lingo, an LLM, a smaller NLP model, rules, retrieval, or a hybrid pipeline. In the prototype, Lingo can realize both broad and narrow linguistic work: \texttt{Engine.create} can extract a typed object, \texttt{Engine.decide} can perform a bounded binary classification, \texttt{Engine.choose} can select from an enumerated set, \texttt{Flow} can encode a procedural workflow, and a \texttt{StateMachine} can constrain a multi-step dialogue.

Generated NLP agents are solution-local. The \texttt{CoordinationAgent} includes their specifications in the \texttt{SolutionProposal}, including purpose, method, expected input, expected output, validation rule, and lifecycle. They do not automatically enter the registry. Promotion to a persistent registered agent would require a separate SDK admission process.

\section{Architecture Diagram}

Figure~\ref{fig:implementation-architecture} summarizes the implementation structure. The diagram is intentionally abstract enough to support later coding choices while still showing that coordination logic belongs to \texttt{CoordinationAgent}, while communication and runtime adaptation belong to \texttt{CoordinationSdk}.

\begin{figure}[htbp]
\centering
\fbox{
\begin{minipage}{0.90\textwidth}
\centering
User Problem or Application Request\\
$\downarrow$\\
\texttt{CoordinationAgent}\\
$\downarrow$\\
\begin{tabular}{ccc}
Problem Interpretation & Feasibility Analyzer & Plan / Result Aggregation \\
\end{tabular}\\
$\downarrow$\\
\texttt{CoordinationSdk} Middleware Boundary\\
$\downarrow$\\
\begin{tabular}{ccc}
A2A Gateway & Local Agent Runtime & Lingo Agent Runtime \\
\end{tabular}\\
$\downarrow$\\
\begin{tabular}{ccc}
Remote A2A Agents & Local Python Agents & Linguistic Agents \\
\end{tabular}\\
$\downarrow$\\
Artifacts, Status, Failures, and Communication Trace\\
$\downarrow$\\
Final Result or Structured Infeasibility Report
\end{minipage}
}
\caption{CoordinationAgent over SDK middleware for A2A-compatible coordination across remote, local, and linguistic agents.}
\label{fig:implementation-architecture}
\end{figure}

\section{Execution Sequence}

The execution sequence begins when the user or application submits a problem request to \texttt{CoordinationAgent.coordinate}. The coordination agent asks the SDK for the current registry snapshot and capability index. It obtains a typed problem representation from a structured caller or by invoking a registered linguistic capability through the SDK. The coordination agent then attempts direct matching before considering decomposition. The \texttt{FeasibilityAnalyzer} checks the candidate plan and either approves it, permits an auxiliary NLP specification to be added, or rejects it as infeasible. Approved tasks are dispatched through \texttt{sdk.send\_task(...)}. The SDK chooses the correct internal runtime, records status and failures, normalizes artifacts, and exposes communication trace events. The coordination agent aggregates the authorized outputs or returns an infeasibility report.

\begin{figure}[htbp]
\centering
\fbox{
\begin{minipage}{0.88\textwidth}
\begin{enumerate}
\item User or application submits a problem and constraints to \texttt{CoordinationAgent}.
\item Coordination agent reads registry and capabilities through \texttt{CoordinationSdk}.
\item Coordination agent obtains or constructs a typed \texttt{ProblemRequest}.
\item Coordination agent attempts direct capability matching before decomposition.
\item Coordination agent requests candidate planning only when direct matching is insufficient.
\item Feasibility Analyzer validates skills, modalities, dependencies, authority, and evidence contracts.
\item Auxiliary Capability Factory adds a temporary specification only when justified.
\item Coordination agent dispatches authorized tasks through \texttt{sdk.send\_task(...)}.
\item SDK records runtime status, errors, artifacts, and communication evidence.
\item Coordination agent returns the final result or a structured infeasibility report.
\end{enumerate}
\end{minipage}
}
\caption{Execution sequence for CoordinationAgent-mediated problem solving.}
\label{fig:execution-sequence}
\end{figure}

\section{Failure and Infeasibility Handling}

The implementation treats infeasibility as a first-class coordination-agent result. If no remote, local, linguistic, or approved auxiliary capability can satisfy a required subtask, the \texttt{CoordinationAgent} returns a structured report. If an agent is unavailable, the coordination agent may select another compatible agent from the SDK registry; if no replacement exists, it reports the missing dependency. If modalities are incompatible, the report identifies which input or output mode cannot be bridged. If a linguistic planner proposes a decomposition that violates explicit constraints, the \texttt{FeasibilityAnalyzer} rejects it.

The SDK normalizes runtime failures rather than planning failures. A2A transport failures, local handler exceptions, Lingo execution failures, invalid artifacts, cancellations, and timeouts are recorded as structured observations linked to the relevant task and agent. This behavior is essential for simplicity and trust. The system should not create complex plans merely to appear capable. It should coordinate only when the available and generated capabilities justify the solution. Otherwise, the coordination agent should explain the boundary clearly. Lingo guardrails may detect obvious unsafe, irrelevant, or ambiguous inputs early, but such reflexes are treated as advisory preprocessing rather than as replacements for the feasibility verdict.

\section{Consensus-Backed Unified Coordinator Nodes}

The current crash-fault-tolerance path preserves one logical authorization role while removing dependence on one coordinator process or one PostgreSQL server. A \texttt{CoordinatorNode} is one container that supervises the Python coordinator service and, when assigned cluster membership, a separate etcd process. The image pins etcd~3.6.13 \cite{etcd3613release}. etcd replicates authoritative registry and coordination state through Raft, whose majority rule provides one committed log order under crash faults and partitions that leave a quorum available \cite{ongaro2014raft}. Co-location reduces deployment units, but the processes remain separate so either can be monitored and terminated explicitly. The coordinator limits concurrent dispatches to reduce the risk that task or model workloads starve Raft heartbeats.

Startup validates node identity, cluster identity, protocol version, shared secrets, advertised API and peer endpoints, voter target, and persistent data directory. Direct and DNS seeds are tried first. If none is reachable, the node sends three jittered authenticated probes over three seconds to the link-local multicast group \texttt{239.255.42.99:7947}; responses are unicast. Discovery, join, leave, and configuration envelopes use canonical JSON and HMAC-SHA256. Timestamps outside 30 seconds, reused nonces, conflicting cluster identifiers, incompatible protocol versions, and invalid signatures fail closed. Multicast discovers an existing cluster; it does not authorize bootstrap. Initial creation still requires exactly one node configured explicitly for bootstrap.

The target number of voting etcd members is stored in authoritative cluster configuration and selected with \texttt{COORDINATION\_VOTER\_TARGET}. Only 1, 3, 5, and 7 are accepted. Three is the default and minimum production recommendation; one is explicitly degraded development or bootstrap mode. An odd target avoids increasing quorum without increasing tolerated failures. The system reports degradation when healthy voters are below target, while coordinator-only nodes may continue to join. This is automatic membership reconfiguration, not automatic provisioning: an external runtime must still create or remove containers.

\begin{table}[htbp]
\centering
\caption{Configurable voter profiles and majority requirements.}
\label{tab:voter-profiles}
\begin{tabular}{rrrrl}
\toprule
\textbf{Target} & \textbf{Quorum} & \textbf{Failures} & \textbf{Mode} & \textbf{Guidance} \\
\midrule
1 & 1 & 0 & degraded & development or bootstrap only \\
3 & 2 & 1 & production & default minimum recommendation \\
5 & 3 & 2 & production & higher availability, higher write cost \\
7 & 4 & 3 & exceptional & use only when failures justify latency \\
\bottomrule
\end{tabular}
\end{table}

Every coordinator runs a \texttt{MembershipReconciler}; an etcd lease elects one active reconciler and another takes over after lease loss. A joining node first records an authenticated intent. If more voters are needed, the reconciler adds it as a learner and promotes it only after etcd confirms synchronization. Otherwise it becomes coordinator-only. Membership changes are serialized one at a time, as recommended for etcd reconfiguration \cite{etcdreconfiguration}. Graceful scale-down drains and removes one voter at a time. A failed voter is replaced only while the remaining members retain quorum and a healthy replacement is available. Short heartbeat loss is insufficient for eviction, and majority loss stops automatic reconfiguration because no safe consensus decision can be committed.

The \texttt{EtcdCoordinationStore} and \texttt{EtcdAgentRegistry} use etcd's v3 HTTP gateway. The client rotates endpoints, bounds retries, performs linearizable reads and transactions, renews leases, and resumes watches from revisions. Session ownership is attached to a lease; the ownership revision is its fencing token. Every state mutation compares owner identity and fencing revision. Attempt creation and its event are atomic, a result requires a started attempt, and terminal creation compares that no terminal key exists. Registry entries use renewable TTL leases, and each authorized plan binds both a registry snapshot hash and its etcd revision. If quorum is unavailable, authorization, dispatch, retry, recovery, and terminal completion fail closed. JSONL and PostgreSQL remain compatibility backends, while PostgreSQL may receive asynchronous audit projections that do not affect readiness.

External effects require an additional receiver-side boundary because consensus cannot make a network side effect exactly once. Dispatch metadata includes an operation key derived from session, plan generation, and task, an attempt key, fencing token, coordinator identity, and registry revision. An effectful agent must durably retain the highest fence observed for an operation and cache completed operation results. Agents that lack this support are restricted to read-only or explicitly idempotent tasks; unsafe or unknown effects are refused. Process-local agents are excluded from fault-tolerant plans unless declared replicated across coordinator nodes.

\begin{table}[htbp]
\centering
\caption{Scope of implemented coordination mechanisms and evidence.}
\label{tab:implementation-scope}
\small
\begin{tabular}{p{0.24\textwidth}p{0.34\textwidth}p{0.30\textwidth}}
\toprule
\textbf{Level} & \textbf{Implemented mechanism} & \textbf{Claim boundary} \\
\midrule
Local prototype & JSONL sessions, deterministic feasibility, SDK dispatch, trace reconstruction, and replay. & Supports simple local execution and compatibility tests. \\
Legacy replication & PostgreSQL leases, fences, idempotency, terminal constraints, and selected crash recovery. & One shared database remains the serialization point. \\
Consensus-backed coordination & Unified nodes, etcd/Raft state, distributed registry, lease-elected reconciliation, learners, odd voter targets, and quorum-fail-closed operations. & Crash-fault tolerant only while a majority survives; not Byzantine or exactly once. \\
Adversarial deployment & Shared-secret admission and private-network assumptions; no mTLS, attestation, or Byzantine coordinator protocol. & Production adversarial resistance remains outside the measured evidence. \\
\bottomrule
\end{tabular}
\end{table}

Operational endpoints separate process liveness from coordination readiness. The liveness endpoint checks the process, while readiness requires reachable quorum, node registration, and working lease operations. Cluster status reports target and active voters, learners, coordinator-only nodes, quorum, leader, revision, discovery path, degradation, and pending changes. Metrics and structured events cover membership, election, quorum, discovery, authentication rejection, stale fencing, registry expiry, recovery, and duplicate suppression.

This design keeps the theory's centralized authorization semantics intact while making its physical state consensus-backed. Several nodes may accept requests, but each session has one fenced owner, one authorized plan per generation, one attempt history, and at most one terminal result. Logical centralization therefore no longer implies a single volatile process, yet safety still depends on trusted coordinator code, a surviving etcd majority, private etcd networking, and receiver-enforced effect controls.

\section{Chapter Synthesis}

The design specifies a distributed multi-agent environment with a clear split between coordination and communication. \texttt{CoordinationAgent} owns planning, feasibility, authorization, auxiliary decisions, and aggregation. \texttt{CoordinationSdk} owns A2A interoperability, local runtime invocation, Lingo runtime invocation, task dispatch, artifact normalization, communication failures, and trace access. Unified coordinator nodes add consensus-backed registry, ownership, and durable state without decentralizing plan authority. Agents remain independently deployable and vendor independent. A2A provides the interoperability backbone through Agent Cards and task exchange. Local Python agents and Lingo-backed linguistic agents are wrapped by the SDK so that they participate through compatible dispatch, artifact, and trace mechanisms; node-local availability is nevertheless excluded from fault-tolerant plans unless replicated. The evaluation in Chapters~\ref{chap:methodology} and~\ref{chap:results} distinguishes the frozen linguistic study from the later distributed implementation checks.
