\chapter{Introduction}
\label{chap:introduction}

\section{Context and Motivation}

Autonomous systems increasingly coordinate components that differ in ownership, implementation, authority, and reasoning method. Large language models (LLMs) extend these systems with natural-language interpretation, flexible task decomposition, and explanation. Symbolic artificial intelligence (AI), in contrast, provides explicit representations of capabilities, constraints, plans, and state transitions. The two approaches are complementary: language models are effective at proposing meaning under ambiguity, whereas symbolic methods are effective at checking conditions that can be stated precisely.

This complementarity does not automatically yield dependable coordination. An LLM can describe a plausible division of work even when no admitted agent possesses a required capability, two tasks exchange incompatible artifacts, or an action exceeds the authority granted to its executor. Free-form dialogue can also obscure whether an agent merely discussed a task, accepted a commitment, or supplied evidence of completion. These distinctions matter whenever coordinated agents can invoke tools, modify shared state, or affect an external environment.

\section{Problem Statement and Research Gap}

Classical multi-agent systems provide explicit interaction protocols, commitments, organizational roles, and formal coordination mechanisms. Recent LLM-agent systems provide broad linguistic flexibility and tool-use interfaces. Emerging interoperability protocols improve connectivity among independently implemented agents. Nevertheless, these lines of work do not by themselves define when a natural-language coordination proposal is sufficiently justified for execution. In particular, connectivity does not establish semantic compatibility, and fluent agreement does not establish feasibility.

The scientific problem addressed by this thesis is therefore:

\begin{quote}
\emph{How can a heterogeneous multi-agent system use LLM-based interpretation and decomposition while ensuring, through explicit symbolic criteria, that an accepted coordination plan is supported by available capabilities, compatible artifacts, applicable authority, and verifiable completion conditions?}
\end{quote}

The problem is not merely the construction of another agent platform. It is the absence of a precise operational boundary between a linguistically plausible proposal and a plan that the current system is justified in authorizing. Existing orchestration frameworks, agent protocols, schemas, validators, and tool registries provide important building blocks; the novelty sought here is the enforced combination of those pieces into a coordination boundary where candidate plans, authority checks, artifact contracts, commitments, refusals, and execution traces are all explicit before dispatch.

\section{Object of Study and Field of Action}

The \emph{object of study} is coordination in heterogeneous autonomous multi-agent systems. The \emph{field of action} is the hybrid mediation of natural-language requests into symbolically constrained plans, commitments, execution traces, and infeasibility decisions for independently implemented agents.

\section{Research Objective}

The general objective is to design and specify a unified coordination framework that uses LLMs for semantic interpretation and candidate decomposition while using symbolic representations and deterministic checks to authorize, trace, or reject multi-agent plans.

The specific objectives are:

\begin{enumerate}
  \item Critically synthesize classical multi-agent coordination, symbolic AI, LLM-agent architectures, and interoperability protocols to identify the unresolved integration gap.
  \item Define a protocol-independent theoretical framework for capabilities, tasks, authority, commitments, artifact compatibility, bounded auxiliary capabilities, and feasibility.
  \item Instantiate the framework as an implementable Python coordination agent over middleware SDK that preserves protocol-mediated interoperability while supporting remote, local, and Lingo-authored agents without coupling the theoretical contribution to a particular model provider, agent framework, or communication protocol.
  \item Evaluate the corrected prototype with an author-labelled, label-hidden, repeated 36-case study, paired controls and ablations, deterministic scenarios, containerized A2A checks, and measured PostgreSQL crash recovery.
  \item Analyze the scope, limitations, validity threats, and practical implications of the proposed framework.
\end{enumerate}

\section{Research Questions}

The thesis is guided by the following questions:

\begin{description}
  \item[RQ1] Which coordination information must be represented explicitly to distinguish an executable multi-agent plan from a linguistically plausible one?
  \item[RQ2] How should responsibility be divided between LLM-based reasoning and symbolic validation in a heterogeneous coordination workflow?
  \item[RQ3] How can missing narrow linguistic capabilities be introduced without granting unbounded new competence or authority?
  \item[RQ4] What do the measured outcome and process results show about justified acceptance, explicit refusal, traceability, dependency discipline, and recovery relative to paired controls and ablations?
\end{description}

\section{Scope and Contributions}

The thesis focuses on plan-level coordination under a trusted central authorization role and distributed execution. It does not claim to verify the internal reasoning of participating agents, solve Byzantine coordination, or guarantee runtime success after authorization. Its contribution is integrative: a formal separation between linguistic proposal and symbolic authorization; typed, fail-closed feasibility and completion contracts; bounded task-local auxiliary capabilities; durable trace and measured crash recovery; a local A2A admission boundary; and a reproducible method for comparing the integration with paired controls. Established component mechanisms, the new operational integration, prototype evidence, and unproven properties are kept distinct.

\section{Methodological Approach}

The research follows a design-science approach. It combines critical literature synthesis, formal and architectural modeling, implementation-oriented specification, and controlled evaluation. The evaluation separates planning feasibility from runtime success, uses pre-registered author labels hidden during inference and repeated local-model runs, tests mechanism ablations, and measures replicated-coordinator ownership through a shared durable store. Chapter~\ref{chap:methodology} explains the executed and gated stages.

\section{Thesis Organization}

Chapter~\ref{chap:state-of-the-art} synthesizes the theoretical foundations and research gap. Chapter~\ref{chap:methodology} defines the research and evaluation method. Chapter~\ref{chap:theoretical-solution} presents the protocol-independent framework. Chapter~\ref{chap:implementation-design} maps it to a coordination-agent architecture over SDK middleware and then selects a concrete interoperability adapter for implementation. Chapter~\ref{chap:results} specifies the evaluation execution and reporting structure. Chapter~\ref{chap:discussion} examines the implications, limitations, and validity threats. The final chapter relates the contributions to the objectives and identifies future work.
