\chapter{Research Methodology}
\label{chap:methodology}

\section{Study Design}

The thesis follows design science: it constructs a protocol-adaptable coordination boundary and evaluates whether the implementation exhibits the declared authorization, refusal, execution, and recovery behavior. The unit of analysis is one coordination attempt, comprising the natural-language request, admitted registry snapshot, typed request, non-authoritative proposal, symbolic predicates, authorization decision, task attempts, artifacts, and persisted trace. Outcome correctness and process correctness are analyzed separately.

The evaluation is staged. Unit and integration tests exercise individual contracts and transitions. Seven deterministic cases exercise the complete in-process workflow. Container harnesses exercise A2A admission and dispatch across processes and lease-backed recovery through PostgreSQL. The primary comparative study is version~0.5: 48 held-out cases arranged as 24 feasible/infeasible matched pairs, three models, five requested seeds, and two independently prompted arms, producing 1,440 primary outputs. Versions~0.3 and~0.4 are retained as historical development evidence rather than the primary comparison.

\section{Versioned, Pre-specified, and Label-hidden Corpus}

The version 0.3 corpus contains 24 base cases, four in each of six strata: direct feasible execution; decomposed feasible execution; feasible planning followed by runtime failure; missing capability or auxiliary overreach; schema, modality, graph, or ordering invalidity; and authority, typed-constraint, availability, or registry invalidity. Twelve controlled variations add four paraphrases, four availability changes, and four capability-alias perturbations. Public requests, registry snapshots, payloads, and shared proposals are stored separately from hidden reference labels.

Version 0.4 was a development cycle created after observing the version 0.3 all-refusal result. It is not retrospectively altered. Because its corpus language and bridge protocol were used to tune the redesign, it is historical rather than confirmatory evidence.

Version 0.5 is fresh and was frozen before collection. Eight domains each contribute three matched pairs, and cyclic offsets assign three pairs to each of eight invalidity categories: missing capability, unavailable agent, insufficient trust, incompatible modality, typed-constraint violation, invalid artifact contract, dependency/order error, and auxiliary overreach. The resulting 48 held-out cases are distinct from eight unscored development examples. Twelve additional runtime cases test dependency failure, timeout and unknown attempts, retry/recovery, artifact rejection, and trace behavior without entering planning accuracy. The protocol is described as \emph{pre-specified}, not preregistered, because it was not deposited in an external timestamped registry.

The thesis author defined the reference labels and minimum justifications before model collection. A provenance record identifies the author, annotation date, labeling protocol, corpus version and SHA-256 digest, and declares the labels frozen. No independent adjudicator was available. Reference labels are never included in model prompts and are loaded by the scorer only after collection is complete. An automated audit checks document shape, unique identities, public/hidden alignment, hash agreement, status consistency, and justification presence; it establishes internal consistency, not independent correctness. Any post-freeze change requires a new corpus version and hash.

\section{Configurations and Fairness}

For every model, seed, and held-out case, two independently prompted arms receive the same public request, admitted registry, payload, and schemas. \texttt{hybrid\_bridge\_v05} emits only canonical requirement, capability, and dependency references; deterministic hydration constructs executable fields and symbolic feasibility makes the authorization decision. \texttt{direct\_llm\_coordinator\_v05} emits an explicit accept/refuse decision and a complete proposed assignment without a plan-level symbolic gate. Its proposal is checked post hoc but is never dispatched. Each arm has the same one-initial-plus-one-repair maximum. Repair feedback is limited to JSON Schema and public referential errors and never reveals labels, failed authority predicates, or hidden expectations.

A non-oracle greedy baseline selects the stable lowest-ranked admitted capability satisfying public predicates and copies request-declared dependencies. The lexical matcher is retained only as a diagnostic, while the structured oracle is explicitly an upper bound. Initial and repaired views are reported for both experimental arms; repair is therefore observable rather than silently folded into the headline result.

Planning-only decisions and runtime compliance are not conflated. Dependency-gating and durable-trace ablations may leave the planning decision unchanged while altering execution safety or observability. Those mechanisms are therefore scored through dispatch, skip, duplicate-effect, and trace-completeness measures rather than through feasibility accuracy alone.

The runtime-only study contains five dependency scenarios, five auxiliary-admission scenarios, and five trace-evidence scenarios, each repeated ten times under the applicable strict and negative-control configurations. This produces 350 deterministic observations in a separate immutable bundle. The dependency control bypasses predecessor status; the auxiliary controls disable the feature or bypass its bounds; and the trace control suppresses returned diagnostic evidence while preserving state-critical persistence. These rows are not treated as LLM planning observations.

\section{Local Model Protocol}

The fixed local models are \path{qwen/qwen3-1.7b}, \path{google/gemma-4-e2b}, and \path{qwen/qwen3-8b}. Qwen 14B is excluded because its installed nine-gigabyte quantization does not leave a reproducible operating margin on the 16~GB host. LM Studio is addressed at \url{http://127.0.0.1:1234/v1} with a non-secret local API-key value. Temperature is zero and seeds are 11, 22, 33, 44, and 55.

All other models are unloaded before a batch. The runner loads one exact identifier and verifies the model catalog. For each of 48 cases, both arms are collected at five requested seeds for each of three models, yielding exactly 1,440 immutable primary outputs. Every record preserves arm/model/seed/case identity, prompt, initial raw response, parsed object, optional repair response, latency, and call count. Repeated identical outputs remain in the evidence set. Interrupted runs may resume only after corpus, protocol-hash, and existing-record identity checks; completed case files are never rewritten and scoring is blocked until the complete matrix exists.

For version 0.4 the complete admitted \texttt{ProblemRequest} and registry snapshot are public prompt inputs, while the output is only a \texttt{LinguisticPlanDraft}. Deterministic hydration occurs after model output. When the draft contains repairable structural errors---such as omitted, duplicate, or unknown requirement references or invalid dependencies---the model receives those public errors once and may return one repaired draft. Capability unavailability and symbolic infeasibility do not trigger repair. Initial linguistic latency, repair latency, symbolic-validation latency, execution latency, and total latency are stored separately.

\section{Metrics and Statistical Analysis}

Decision measures include accuracy, precision, recall, F1, false acceptance, false refusal, status agreement, and category-level errors. Boundary measures include constraint, authority, schema, modality, dependency, and evidence rejection; artifact-contract validity; required-artifact completion; authorization-before-dispatch; dependent-task skipping; capability-ID resolution; unresolved references; and trace completeness and isolation. Efficiency measures include messages, planner calls, tokens, median and interquartile latency, and coordination overhead. Recovery measures include duplicate dispatch, repeated effect, stale-write rejection, recovery latency, and terminal correctness.

The 24 matched pairs are the primary statistical units; model and requested-seed outputs are repeated measurements. The analysis reports initial-to-repaired change, repair rate and success by model/domain/invalidity, repeated-observation and case-majority metrics, and model/seed stability without treating requested seeds as independent samples. Exact paired McNemar tests compare repaired arms within each model and use Holm correction. A matched-pair clustered bootstrap reports a 95\% interval for the paired accuracy difference. Functional results are deterministic and hashable; initial call, repair, hydration, symbolic validation, post-hoc validation, and total wall-clock timings are stored separately as hardware-dependent evidence.

The pre-specified criteria are zero measured hybrid false acceptance; at least 60\% feasible recall for every model; a positive paired case-level accuracy difference between repaired hybrid and repaired direct LLM, with the interval and exact tests reported whether significant or not; and 100\% authorization-before-dispatch and trace-order compliance in the separate runtime suite. The first complete clean matrix is reported non-selectively even if a criterion fails.

\section{Consensus Crash-Fault Campaign}

RQ5 is evaluated as a separate core track rather than as additional LLM observations. Before accepted execution, the campaign protocol fixes three fresh trials for clean formation at voter targets three, five, and seven; live transitions $3\rightarrow5\rightarrow3$ and $5\rightarrow7\rightarrow5$; leader termination; minority partition; majority loss and restoration; persistent restart; failed-voter replacement; unavailable audit projection; concurrent ownership; and crashes after attempt start, after external dispatch, and during aggregation. Every trial uses a new Docker Compose project, new volumes except where persistence is the object of the test, and a non-existing evidence directory.

Safety observations are one committed session owner, monotonic fencing, rejection of stale writes, attempt start before completion, at most one terminal result, refusal of authoritative mutation without quorum, serialized membership changes, and zero repeated external effects when the receiving fixture enforces operation keys and fences. Conditional liveness requires leader-failure progress within 30 seconds while a majority survives, quorum-loss detection within 10 seconds, restored progress within 30 seconds, and membership convergence within 120 seconds. Steady state additionally requires authoritative role agreement and no pending membership operation. Repetition tests implementation stability but does not create independent deployment populations.

The campaign invokes etcd~3.6.13 as the Raft implementation; it does not reimplement or re-prove Raft. The fault model is crash/omission under eventual synchrony with a surviving majority. All containers share one Docker host, so the campaign does not represent independent machines, disks, regions, switches, or Byzantine behavior.

\section{Provenance and Reproducibility}

Each run directory records Git SHA, dirty-state flag, corpus hash, model ID, prompt/schema version, seed, dependency-lock hash, Docker image digest, Python, PostgreSQL, A2A and LM Studio versions, operating system, CPU, RAM, and available accelerator information. Reports are written to new run-ID directories; convenience summaries are derived and may be regenerated, but the only copy of a raw run is never overwritten.

The release gates require all unit and integration tests, static lint and type checks, the seven deterministic cases, eleven Docker A2A checks, five PostgreSQL checks, the upstream-derived A2A check, the full consensus campaign, evidence-schema checks, and LuaLaTeX/BibTeX compilation. Coverage thresholds are 85\% overall production code; 90\% for feasibility, coordination execution, trace validation, plan hydration, and evidence preflight; 85\% for cluster discovery, launcher, etcd, registry, service, and SDK modules; and 80\% for PostgreSQL, the consensus campaign driver, and current evaluation modules. Fixture servers and system drivers are exercised as subprocesses or containers rather than counted in the production aggregate. A release cannot be frozen while any threshold, clean-source evidence, institutional metadata, evaluated-source commit, or packaging-commit field is open.

\section{Validity Boundaries}

The corpus is designed and labelled by the thesis author rather than sampled from a deployment population or independently adjudicated. This creates researcher-bias, construct-circularity, and baseline-design threats: agreement measures conformance to the declared framework criteria, not neutral benchmark truth. Freezing labels before collection and hiding them from prompts reduces leakage but does not create independence. The agents are mostly controlled fixtures, and the local models are small. Contract validity does not establish semantic truth. The PostgreSQL lease-and-fencing evidence concerns measured crash faults and is not consensus evidence. The etcd campaign is a core implementation evaluation but remains a single-host, project-authored fault campaign rather than proof of all Raft properties or multi-host availability. A2A admission establishes local origin, transport, trust, and credential policy; it does not establish cryptographic identity or malicious-agent resistance. These boundaries constrain every answer to the research questions.
