\chapter{Evaluation and Results}
\label{chap:results}

This chapter reports the empirical and implementation evidence used to answer the research questions. Its goal is not to turn reproducibility into a positive result: it distinguishes evidence validity from claim support and therefore reports failed criteria, unsafe ablations, infrastructure failures, and unexecuted checks alongside successful tests.

The chapter first establishes the status and provenance of every evidence family. It then reports the fail-closed boundary, dependency and recovery behavior, end-to-end harnesses, and consensus campaign. Runtime-only ablations are kept separate from the comparative planning study. The final three sections preserve the historical v0.3 and v0.4 findings before presenting v0.5 as the primary comparison, including the direct-LLM control, repair behavior, statistical uncertainty, and the negative outcome for the proposed comparative advantage.

\section{Evidence Status}

The evidence base distinguishes historical version 0.2--0.4 artifacts from the accepted version~0.5 primary comparison and two clean 33-trial consensus campaigns. Earlier reports remain development artifacts and are excluded from the final RQ4 comparison. The v0.5 corpus is author-labelled, pre-specified, frozen, and label-hidden during inference; no independent adjudicator was available. Evidence validity is separated from outcome: the v0.5 result is valid but mixed, and both complete consensus campaigns are valid but failed.

\begin{table}[htbp]
\centering
\caption{Repository evidence status after the distributed implementation update.}
\label{tab:evidence-status}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.44\textwidth}>{\raggedright\arraybackslash}p{0.18\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}}
\toprule
\textbf{Gate} & \textbf{Observed} & \textbf{Required} \\
\midrule
Python tests & \EvidenceTestCount{} passing & all passing \\
Overall production branch coverage & \ProductionBranchCoverage\% & at least 85\% \\
Deterministic scenarios & 7/7 & 7/7 \\
Docker A2A checks & 11/11 & 11/11 \\
PostgreSQL distributed checks & 5/5 & 5/5 \\
Historical etcd three-voter smoke checks & 7/7 & historical only \\
Full 3/5/7 consensus campaign & \ConsensusTrialCount/33; \ConsensusOutcome & 33 accounted trials; criteria reported \\
Upstream-derived A2A interoperability & 1/1 & passing \\
Ruff and mypy & passing & passing \\
Frozen author-label provenance & complete & complete \\
Historical v0.4 outputs & 540/540 & historical only \\
Primary v0.5 outputs & \VFiveOutputCount/1,440 & 1,440/1,440 \\
Deterministic schema 0.5.0 analysis & reproducible & hash-reproducible \\
\bottomrule
\end{tabular}
\end{table}

Branch-aware production coverage is \ProductionBranchCoverage\%, above the 85\% aggregate gate. Every independently enforced critical-module threshold passes, including at least 85\% for cluster discovery, at least 80\% for the consensus campaign driver, and at least 90\% for evidence preflight. All \EvidenceTestCount{} tests, Ruff, mypy, the production aggregate, and the module-specific gates pass on the recorded Windows run. The generated thesis macros are derived from the manifest and reproduce without a diff.

\section{Fail-Closed Symbolic Boundary}

Regression tests show that a promised artifact does not make a validatorless task verifiable; legacy textual constraints become unresolved typed constraints and force refusal; missing JSON-pointer paths, false predicates, canonical capability-ID mismatch, incomplete mode coverage, and nested JSON Schema type violations are rejected. Effective trust is assigned by local admission policy rather than copied from a remote Agent Card. Cards requiring an unavailable credential scheme are refused, and plain HTTP requires an explicit development exception.

Candidate plans remain non-authoritative. Direct delegation is selected only if it is feasible; otherwise bounded candidates are checked and ranked deterministically by task count, auxiliary count, DAG depth, and stable task identifiers. These tests support prototype correctness of the implemented predicates, not mathematical proof of arbitrary plan correctness.

\section{Dependency, Recovery, and Trace Behavior}

Execution tests confirm that a task is dispatched only after every declared prerequisite completes. Failure, timeout, refusal, or unknown outcome skips descendants while independent branches continue. Consumers receive artifacts only from declared dependencies. On takeover, unsafe unknown work is recorded as \texttt{unknown} and is not blindly redispatched. Completion additionally requires the typed completion contract.

Concurrent-session tests show that returned traces contain only persisted events for their session. Authorization precedes attempt start, SDK transport observations are persisted with session, plan, task, and attempt identity, and a trace validator checks ordering and terminal observations. Registration diagnostics remain outside operational session traces.

\section{Corrected End-to-End Harnesses}

The seven deterministic statuses match their references: three successful completions, one authorized runtime failure, and three pre-dispatch infeasibility decisions. The privacy case now derives email and phone values from transcript text; the answers are no longer supplied as separate fixture fields. The task-local auxiliary lifecycle expires after use.

The corrected Docker harness passes eleven checks covering health, registry discovery, direct and decomposed A2A execution, missing capability, insufficient authority, incompatible modes, invalid artifacts, timeout, registry outage, and session replay. The decomposed case originally exposed a stale assumption that every prior artifact should be forwarded. The corrected planner declares dependencies from canonical mode flow, and the harness now verifies that the final consumer receives only its declared predecessor artifact.

The first clean PostgreSQL harness failed because lease acquisition used a read-then-upsert sequence that allowed a competing coordinator to reach a duplicate terminal insertion. That report is preserved. Commit \texttt{3f8093af1ecc7c49312ac34856bba825bd81f381} replaced acquisition with one conditional PostgreSQL upsert that refuses a live foreign holder and increments fencing atomically for an expired or same-holder lease. The subsequent clean harness passes all five checks: live lease conflict, takeover after an abandoned read-only attempt, process exit after external dispatch of unknown-side-effect work, stale-fence rejection, and database invariant rejection. It records zero duplicate dispatches and repeated receiver-side effects, one lease conflict, one stale-write rejection, and terminal correctness. These observations do not establish consensus or cover all crash windows.

An additional test launches the vendored, upstream-derived A2A Hello World sample as an independent server, ingests its Agent Card, preserves its \texttt{echo\_bot} skill identifier, authorizes the matching task, and receives its remote artifact. Its provenance records the upstream URL, commit, blob identifiers, copied-file hashes, license, and local comment-only adaptation. This establishes one reproducible interoperability path; it is not evidence of universal A2A compatibility.

\section{Consensus-Backed Distributed Evidence}

The primary clean campaign was built once from correction commit \texttt{3f8093af1ecc7c49\allowbreak 312ac34856bba825\allowbreak bd81f381} as image \texttt{sha256:573293cdc992ce04\allowbreak 94986e70c25dd8cf\allowbreak 3ba59e1ae52c9ce2\allowbreak 6a4dc16e54eca410}; every isolated trial then used \texttt{--no-build}. It completed the exact 33-trial matrix with complete check accounting. Evidence validity is true although the outcome is failed and the claim status is unsupported. Of 33 trials, 20 passed, seven violated invariants, and six ended in infrastructure errors. Only \ConsensusChecksExecuted{} of \ConsensusChecksExpected{} expected checks executed; 19 executed checks violated their invariant and 40 remained unexecuted. It would therefore be false to summarize the campaign as ``zero safety violations.''

\begin{table}[htbp]
\centering
\caption{Primary clean consensus campaign outcomes across three repetitions.}
\label{tab:consensus-campaign}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.47\textwidth}rrr}
\toprule
\textbf{Scenario} & \textbf{Pass} & \textbf{Invariant fail} & \textbf{Infra. error} \\
\midrule
Formation at 3, 5, and 7 voters (combined) & 9 & 0 & 0 \\
$3\rightarrow5\rightarrow3$ reconfiguration & 3 & 0 & 0 \\
$5\rightarrow7\rightarrow5$ reconfiguration & 1 & 0 & 2 \\
Leader/partition/quorum/concurrency & 0 & 0 & 3 \\
Audit sink unavailable & 3 & 0 & 0 \\
Failed-voter replacement & 1 & 1 & 1 \\
Crash after attempt start & 0 & 3 & 0 \\
Crash after external dispatch & 0 & 3 & 0 \\
Crash during aggregation & 3 & 0 & 0 \\
\bottomrule
\end{tabular}
\end{table}

A prior complete clean campaign from commit \texttt{7a31b42069104c3a\allowbreak 4d7f10d8441d8235\allowbreak 7a35fd8b} is also preserved: 19 trials passed, eight violated invariants, and six had infrastructure errors, with 102/144 checks executed and 20 violations. The similar but non-identical distributions demonstrate both reproducible fault classes and run-to-run instability. Formation, audit failure, and aggregation-crash cases were consistently successful. Reconfiguration could leave a pending membership change, the combined scenario repeatedly failed before executing its checks, and the attempt-start and post-dispatch crash windows repeatedly violated duplicate-effect or recovery invariants. Consequently, RQ5 receives a negative empirical answer for this implementation and single-host campaign. The mechanism exists, but the measured system does not preserve all stated safety and conditional-liveness properties.

\section{Separate Runtime-Only Ablations}

The runtime bundle contains 350 deterministic observations: 100 dependency-dispatch observations, 150 auxiliary-admission observations, and 100 trace-evidence observations. The secure defaults produced zero safety violations across their 150 applicable observations. Removing dependency gating produced 40 unsafe dispatch decisions in the 40 predecessor-missing, failed, timed-out, or unknown observations. Removing auxiliary bounds admitted 40 of 40 deliberately ineligible, persistent, unscoped, or write-authorized specifications; disabling auxiliary behavior admitted none, including ten otherwise valid bounded cases. Suppressing durable diagnostic trace evidence removed the required observation in all 50 trace cases while leaving state-critical persistence unchanged. The observation-file SHA-256 is \texttt{7617bb8da5308b9a\allowbreak 22a7ed7fe12c93d7\allowbreak e0b80484b2a18de2\allowbreak ee44ea3d2e3b7bbf}.

These controls isolate mechanism consequences rather than planning quality. Repetition confirms deterministic behavior of the implemented policy surfaces but does not create 350 independent scenarios: the scenario types, not repeated executions, bound external validity. In particular, the trace result concerns availability of auditable returned evidence, not deletion of leases, fencing tokens, attempts, or terminal results.

\section{Comparative Conformance Study}

The public corpus contains 36 cases and has hash \texttt{b9fd8b6f8398eaf3\allowbreak 3b8676e14288f0d3\allowbreak c4956cca5a9b915c\allowbreak a7b68990167c9c1d}. Its provenance declares author annotation, corpus freeze, and the absence of independent adjudication. The collector verified this exact hash and never included reference labels in prompts. The scorer verified all 540 unique model--seed--case identities before loading labels. Two analysis reruns produced the same SHA-256 digest, \texttt{b70a169d8527b2f4\allowbreak b95a5ef1a7ac083e\allowbreak 0eb81874c896f504\allowbreak 12352b37b14b03a3}. These are leakage, completeness, and reproducibility controls, not evidence that the labels are independently correct.

\begin{table}[htbp]
\centering
\caption{Decision results on 540 repeated observations; intervals are clustered by the 36 cases.}
\label{tab:v03-comparison}
\begin{tabular}{p{0.35\textwidth}rrrr}
\toprule
\textbf{Configuration} & \textbf{Accuracy} & \textbf{95\% CI} & \textbf{FA} & \textbf{FR} \\
\midrule
Hybrid & 58.3\% & 41.7--75.0\% & 0 & 225 \\
Natural-language rule-only & 58.3\% & 41.7--75.0\% & 0 & 225 \\
Structured oracle upper bound & 100.0\% & 100.0--100.0\% & 0 & 0 \\
LLM-only without symbolic gate & 43.9\% & 28.5--59.3\% & 303 & 0 \\
Each planning ablation & 58.3\% & 41.7--75.0\% & 0 & 225 \\
\bottomrule
\end{tabular}
\end{table}

The hybrid refused every feasible observation: precision, recall, and F1 for feasible acceptance were all zero. Its 58.3\% accuracy therefore reflects correct refusal of infeasible cases rather than useful end-to-end acceptance. On feasible observations, failed predicates most often concerned coverage (225), total assignment (191), completion (185), verifiability (147), and ordering (86). The recurring risks show that local-model proposals omitted accountable executors, canonical requirement identities, promised artifacts, completion contracts, or deterministic validation evidence. The symbolic gate behaved fail-closed, but the linguistic-to-typed-plan bridge did not produce executable plans reliably.

The LLM-only configuration achieved 100\% recall but admitted 303 infeasible repeated observations; its feasible-class precision was 42.6\% and F1 was 59.8\%. The structured oracle achieved 100\% on the author-declared criteria, showing that the symbolic evaluator can accept and refuse correctly when supplied the intended structured request and proposal. At the primary 36-case unit, the oracle differed from the hybrid with Holm-adjusted exact McNemar $p=0.000366$; the LLM-only difference was not significant after correction ($p_{\mathrm{Holm}}=1.0$). Rule-only and the three planning ablations had no discordant decisions relative to the hybrid. This floor effect makes those ablations uninformative rather than evidence that their mechanisms do not matter.

Median collected linguistic latency was 4.84~s overall: 3.61~s for Qwen3 1.7B, 4.66~s for Gemma 4 E2B, and 27.39~s for Qwen3 8B. Configurations share these raw outputs, so this latency is not a between-configuration performance comparison. Symbolic-validation and execution latency belong to the separate controlled runtime evidence.

RQ1--RQ3 receive implementation-level support from the corrected boundary and harnesses. The version 0.3 evaluation answers RQ4 negatively for that protocol: symbolic refusal prevented unsafe acceptance, but model-generated plans made the hybrid unusably conservative. This frozen result supports the authorization boundary while falsifying any claim that the original linguistic integration provided effective accepted coordination. It remains limited to the author's declared criteria on 36 designed cases and does not establish general superiority.

\section{Version 0.4 Linguistic-Bridge Redesign}

Version 0.4 preserves version 0.3 as negative evidence and evaluates a new corpus/protocol hash, \texttt{a0ecd2bf0ebc6740\allowbreak a205108e45385eb1\allowbreak e2e7d7ae00dacab3\allowbreak 8bc123c8ca49aa2d}. The model generated only canonical requirement/capability selections and requirement-level dependencies; deterministic hydration supplied executable fields from the admitted request and registry. One bounded structural repair was attempted in 424 of 540 primary outputs. Analysis schema 0.4.1 scores deterministic controls once per case, uses the initial response for LLM-only scoring, and attributes latency to the work performed by each configuration. Two reruns using the calibrated timing bundle produced the same SHA-256 digest, \texttt{16a1faf6b3fff50a\allowbreak ffb8c2a07b9351f3\allowbreak 7dc90d06a8e76640\allowbreak 3b8defbdad96c527}. The raw run records a dirty source tree and is consequently retained only as historical development evidence; v0.5 supersedes it as the primary comparison.

\begin{table}[htbp]
\centering
\caption{Version 0.4.1 bridge results; LLM configurations have 540 repeated observations, while deterministic controls have 36 case observations.}
\label{tab:v04-comparison}
\begin{tabular}{p{0.37\textwidth}rrrr}
\toprule
\textbf{Configuration} & \textbf{Accuracy} & \textbf{Recall} & \textbf{FA} & \textbf{FR} \\
\midrule
Hybrid constrained bridge & 90.0\% & 76.0\% & 0 & 54 \\
Natural-language rule-only ($n=36$) & 58.3\% & 0.0\% & 0 & 15 \\
Structured oracle upper bound & 100.0\% & 100.0\% & 0 & 0 \\
LLM-only without symbolic gate & 55.4\% & 22.2\% & 66 & 175 \\
Typed-constraint ablation & 89.8\% & 76.0\% & 1 & 54 \\
Authority ablation & 89.8\% & 76.0\% & 1 & 54 \\
Contract-validation ablation & 90.0\% & 76.0\% & 0 & 54 \\
\bottomrule
\end{tabular}
\end{table}

The hybrid's 95\% case-clustered accuracy interval is 85.0--94.6\%. At the case-unit level, all 15 feasible cases had majority acceptance. Nevertheless, only two of three pre-registered success criteria passed. Qwen3 1.7B achieved 28.0\% feasible recall and 70.0\% accuracy; Gemma E2B and Qwen3 8B each achieved 100\% recall and accuracy. Consequently the required 60\% per-model recall floor failed, and version 0.4 is a substantial improvement rather than a fully successful pre-registered validation. The changed corpus wording and bridge protocol also prevent treating the difference from version 0.3 as a direct paired causal estimate.

The corrected latency accounting reports median total latency of approximately 1.91~s for the hybrid and 1.07~s for the initial-response LLM-only configuration; the deterministic lexical and oracle controls require only their local matching or symbolic work. Requested-seed outputs are reported as stability repetitions because the backend response does not confirm seed compliance. Across the five requested seeds, many model--case outputs are identical, so repetition is not interpreted as an increase beyond 36 primary case units.

\section{Version 0.5 Primary Comparison}

The clean v0.5 run contains all 1,440 independently prompted arm--model--seed--case outputs. Labels remained inaccessible until both arms completed, each arm received the same public inputs and at most one structural repair, and two analyzer runs produced the same functional digest, \texttt{5bead15a76a84f66671b9bc14dd7ac6ead9c9a5ce4b434757d94a5e86c093f82}. Wall-clock timings are stored separately. Requested seeds measure stability and are not treated as independent case samples.

\begin{table}[htbp]
\centering
\caption{Primary v0.5 planning results over 720 repeated observations per arm.}
\label{tab:v05-comparison}
\begin{tabular}{p{0.38\textwidth}rrrr}
\toprule
\textbf{Configuration} & \textbf{Accuracy} & \textbf{Recall} & \textbf{FA} & \textbf{FR} \\
\midrule
Hybrid, initial & 50.00\% & 0.00\% & 0 & 360 \\
Hybrid, repaired & 50.69\% & 1.39\% & 0 & 355 \\
Direct LLM, initial & 52.08\% & 100.00\% & 345 & 0 \\
Direct LLM, repaired & 52.08\% & 100.00\% & 345 & 0 \\
Greedy symbolic baseline ($n=48$) & 100.00\% & 100.00\% & 0 & 0 \\
Lexical diagnostic ($n=48$) & 50.00\% & 0.00\% & 0 & 24 \\
Structured oracle ($n=48$) & 100.00\% & 100.00\% & 0 & 0 \\
\bottomrule
\end{tabular}
\end{table}

Repair was invoked for every output in both arms. It changed the direct arm's decision in 455 responses, but those changes were ignored by the frozen decision contract; it recovered only five hybrid feasible observations. By model, repaired hybrid recall was 0\% for Qwen3 1.7B, 4.17\% for Gemma E2B, and 0\% for Qwen3 8B. Thus the zero-false-acceptance criterion passed, but the 60\% per-model feasible-recall criterion failed for every model.

The repaired hybrid was 1.39 percentage points less accurate than the repaired direct coordinator. The matched-pair clustered bootstrap 95\% interval was $[-3.47,0.00]$ percentage points, and within-model exact McNemar comparisons remained non-significant after Holm correction. Case-majority accuracy was 50\% for both arms. The direct arm achieved acceptance utility only by accepting nearly every case, including 345 unsafe repeated observations; the hybrid achieved fail-closed behavior only by refusing nearly every feasible case. Therefore v0.5 supports the safety behavior of the authorization boundary but not useful hybrid coordination or comparative superiority. RQ4 is partially supported in this narrow sense and otherwise answered negatively.
